---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Describe** *maximum likelihood estimation* for linear regression 
* **Describe** *hypothesis testing* for linear regression coefficients, including the sampling distribution used and degrees of freedom

# Estimation

## What is estimation?

:::: {.columns}

::: {.column width="50%"}
* Method by which we get **estimates** of regression coefficients 
* What are the $b_0$ and $b_1$ for the regression equation?
:::

::: {.column width="50%"}

```{r}
#| code-fold: true
height <- 3*rnorm(50) + 70
weight <- 7 * (height - 70) + 25 * rnorm(50) + 125
dat <- data.frame(height, weight)
ggplot(data = dat, aes(x = height, y = weight)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

::::

## Ordinary least squares (OLS) estimation

* **Minimize** the sum of **squared** residuals
* Find $b_0$ and $b_1$ that give the smallest $\Sigma((Y_i - \hat{Y}_1)^2)$
* For linear regression, there is **one** value of $b_0$ and **one** value of $b_1$ that minimize the sum of squared residuals
  * Not true for all estimation methods

## Minimizing

:::: {.columns}

::: {.column width="50%"}
* Functions with squared in them look like a "U"
* Minimum is the bottom of the "U"
* **Tangent line** is a line that touches a curve *at a single point*
  * The **slope** of the tangent line is 0 at the **minimum**

:::

::: {.column width="50%"}

* $Y = X^2$

```{r}
#| code-fold: true
x <- c(-10:10)
y <- x*x
dat1 <- data.frame(x, y)
ggplot(data = dat1, aes(x = x, y = y)) + 
  geom_line(linewidth = 1)
```

:::

::::

## Tangent lines

```{r}
#| code-fold: true
b1 <- c(0:20)
ssr <- b1*b1 - 20*b1 + 115
dat_ssr <- data.frame(b1, ssr)
ggplot(data = dat_ssr, aes(x = b1, y = ssr)) +
  geom_line(linewidth = 1) +
  ylim(0, 120) +
  labs(x = "Regression coefficient (b)", y = "Sum of squared residuals")
```

## Tangent lines (at b = 15)

```{r}
#| code-fold: true
ggplot(data = dat_ssr, aes(x = b1, y = ssr)) +
  geom_line(linewidth = 1) +
  ylim(0, 120) +
  labs(x = "Regression coefficient (b)", y = "Sum of squared residuals") +
  geom_abline(intercept = -110, slope = 10, linewidth = 1, linetype = "dashed", color = "red") +
  annotate("pointrange", x = 15, y = 15*15 - 20*15 + 115, ymin = 15*15 - 20*15 + 115, ymax = 15*15 - 20*15 + 115) +
  geom_segment(x = 15, xend = 15, y = -1, yend = 15*15 - 20*15 + 115, linetype = "dotted", linewidth = 1)
```

## Tangent lines (at b = 12)

```{r}
#| code-fold: true
ggplot(data = dat_ssr, aes(x = b1, y = ssr)) +
  geom_line(linewidth = 1) +
  ylim(0, 120) +
  labs(x = "Regression coefficient (b)", y = "Sum of squared residuals") +
  geom_abline(intercept = -29, slope = 4, linewidth = 1, linetype = "dashed", color = "red") +
  annotate("pointrange", x = 12, y = 12*12 - 20*12 + 115, ymin = 12*12 - 20*12 + 115, ymax = 12*12 - 20*12 + 115) +
  geom_segment(x = 12, xend = 12, y = -1, yend = 12*12 - 20*12 + 115, linetype = "dotted", linewidth = 1)
```

## Tangent lines (at b = 10)

```{r}
#| code-fold: true
ggplot(data = dat_ssr, aes(x = b1, y = ssr)) +
  geom_line(linewidth = 1) +
  ylim(0, 120) +
  labs(x = "Regression coefficient (b)", y = "Sum of squared residuals") +
  geom_abline(intercept = 15, slope = 0, linewidth = 1, linetype = "dashed", color = "red") +
  annotate("pointrange", x = 10, y = 10*10 - 20*10 + 115, ymin = 10*10 - 20*10 + 115, ymax = 10*10 - 20*10 + 115) +
  geom_segment(x = 10, xend = 10, y = -1, yend = 10*10 - 20*10 + 115, linetype = "dotted", linewidth = 1)
```

## Tangents and minimums

* The tangent line is **horizontal** (slope = 0) at the **minimum**
* Want to find the **minimum** of the *sum of squared residuals*
* **Value of regression coefficient** that meets the *least squares criterion*
* Use calculus: **Derivative** of a function produces the *tangent line*

## Least squares solution

1. State **function** to minimize: $\Sigma((Y_i - \hat{Y}_i)^2)$
2. Take **derivative** of the function, with respect to the *constants of interest* (i.e., $b_0$ and $b_1$)
3. Set the derivative **equal to 0**: Normal equation
4. **Solve** for the *constants*

## 1. Function to minimize

$$\Sigma(Y_i - \hat{Y}_i)^2 = $$

$$\Sigma(Y - (b_1 X + b_0))^2 =$$
$$\Sigma(Y - b_1 X - b_0)^2 =$$
$$\Sigma(Y^2 + {b^2_0} + {b^2_1} X^2 - 2 b_0 Y - 2 b_1 X Y + 2 b_0 b_1 X ) =$$
$$\Sigma Y^2 + \Sigma{b^2_0} + \Sigma{b^2_1} X^2 - \Sigma 2 b_0 Y - \Sigma 2 b_1 X Y + \Sigma 2 b_0 b_1 X  =$$
$$\Sigma Y^2 + n{b^2_0} + {b^2_1} \Sigma X^2 - 2 b_0 \Sigma Y - 2 b_1 \Sigma X Y + 2 b_0 b_1 \Sigma X$$

## 2. Get the derivative

* With respect to $b_1$:

$$\frac{\partial \Sigma (Y - \hat{Y})^2}{\partial b_1} = 2 b_1 \Sigma X^2 - 2 \Sigma X Y + 2 b_0 \Sigma X$$

* With respect to $b_0$:

$$\frac{\partial \Sigma (Y - \hat{Y})^2}{\partial b_0} = 2 n b_0 - 2 \Sigma Y + 2 b_1 \Sigma X$$

## 3 and 4. Set to 0 and solve

For $b_1$: 

$$2 b_1 \Sigma X^2 - 2 \Sigma X Y + 2 b_0 \Sigma X = 0 $$
$$ \vdots $$
$$b_1 = \frac{n \Sigma X Y - (\Sigma X) (\Sigma Y)}{n \Sigma X^2 - (\Sigma X)^2} = \frac{SP_{XY}}{SS_X} = \frac{s_{XY}}{{s^2_X}}$$

## 3 and 4. Set to 0 and solve

For $b_0$:

$$2 n b_0 - 2 \Sigma Y + 2 b_1 \Sigma X = 0$$
$$ \vdots $$
$$b_0 = \overline{Y} - b_1 \overline{X}$$

## How one variable **varies**

* Variation = **sum** of **squared deviations** from the **mean**
  * $SS_X = \Sigma((X - \bar{X})^2)$ 
* Variance = **variation** divided by $n - 1$
  * $s^2_X = \frac{SS_X}{n-1}$
* Standard deviation = **square root** of **variance**
  * $s_X = \sqrt{s^2_X}$
  
## How two variables **covary**

* Covariation = **sum** of **product of deviations** from the **mean**
  * $SP_{XY} = \Sigma((X - \bar{X})(Y - \bar{Y}))$ 
* Covariance = **covariation** divided by $n - 1$
  * $s_{XY} = \frac{SP_{XY}}{n-1}$

## Least squares estimates

* $b_1$ is the 
  * **Covariation** between X and Y, divided by the **variation** of X
  * **Covariance** between X and Y, divided by the **variance** of X

$$b_1 = \frac{SP_{XY}}{SS_X} = \frac{s_{XY}}{{s^2_X}}$$

$$b_0 = \overline{Y} - b_1 \overline{X}$$

## Aside: $R^2_{multiple}$

* $R^2$ = proportion of variation in $Y$ that's explained by the predictors
  * Also: Maximum possible squared correlation between $Y$ and $\hat{Y}$

$$R^2_{multiple} = r^2_{Y\hat{Y}} = \frac{SS_{regression}}{SS_Y} = \frac{predictable\ variation}{total\ variation}$$

where $SS_{regression} = \Sigma(\hat{Y_i} - \overline{Y}_i)^2$

## How do we really estimate?

* Maximum likelihood (ML) 
  * **Maximizes** a likelihood function 
  * Makes **assumptions** about the distribution of parameter 

## Likelihood

* **Conceptually** similar to *probability*
  * Higher values are more likely / probable
  * Likelihood = **relative** probability
* Make **assumptions** about the distribution of observations
  * For example, that they're **normally distributed**
  * Use (normal) **probability distributions** to create a likelihood function

## Recall: Normal distribution

$$f(X) = \frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(X - \mu)^2}{2\sigma^2}}$$

* $\mu$ = mean = location of the distribution
* $\sigma^2$ = variance = spread of the distribution
  * $\sigma$ = standard deviation
  
## Likelihood for one subject

* Make **assumptions** about the **population distribution**
* Assume outcome (BDI) has mean = 20, SD = 5
* For a single subject with BDI = 21:
  * $L_i = \frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(X - \mu)^2}{2\sigma^2}} = \frac{1}{\sqrt{2\pi*5^2}}e^{-\frac{(21 - 20)^2}{2*5^2}} = 0.078$
* $L_i$ = 0.078 is the **likelihood** or **relative probability** of X = 21 from a normally distributed population with mean = 20 and SD = 5

## Subject likelihoods

* Small example from Enders (2005)
* Assume $\mu$ = 20, $\sigma$ = 5

```{r}
#| code-fold: true
ID <- c(1:10)
BDI <- c(5, 33, 17, 21, 13, 17, 5, 13, 17, 13)
Likelihood <- dnorm(BDI, 20, 5)
LogLikelihood <- log(Likelihood)
LL_dat <- data.frame(ID, BDI, Likelihood, LogLikelihood)
round(LL_dat, 3)
```

## Normal distribution, mean = 20, SD = 5

```{r}
#| code-fold: true
norm_plot <- ggplot(data = data.frame(x = c(0, 40)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 20, sd = 5)) + 
  ylab("Likelihood")
norm_plot + 
  annotate("pointrange", x = LL_dat[4,2], y = LL_dat[4,3], ymin = LL_dat[4,3], ymax = LL_dat[4,3], size = 1) +
  annotate("text", x = LL_dat[4,2] + 2, y = LL_dat[4,3], label = "0.078", hjust = 0, size = 8) +
  geom_segment(x = LL_dat[4,2], xend = LL_dat[4,2], y = 0, yend = LL_dat[4,3], linetype = "dashed") + 
  annotate("pointrange", x = LL_dat[2,2], y = LL_dat[2,3], ymin = LL_dat[2,3], ymax = LL_dat[2,3], size = 1) +
  annotate("text", x = LL_dat[2,2] - 2, y = LL_dat[2,3], label = "0.003", hjust = 1, size = 8) +
  geom_segment(x = LL_dat[2,2], xend = LL_dat[2,2], y = 0, yend = LL_dat[2,3], linetype = "dashed") + 
  annotate("pointrange", x = LL_dat[3,2], y = LL_dat[3,3], ymin = LL_dat[3,3], ymax = LL_dat[3,3], size = 1) +
  annotate("text", x = LL_dat[3,2] - 2, y = LL_dat[3,3], label = "0.067", hjust = 1, size = 8) +
  geom_segment(x = LL_dat[3,2], xend = LL_dat[3,2], y = 0, yend = LL_dat[3,3], linetype = "dashed") + 
  annotate("pointrange", x = LL_dat[1,2], y = LL_dat[1,3], ymin = LL_dat[1,3], ymax = LL_dat[1,3], size = 1) +
  annotate("text", x = LL_dat[1,2] + 2, y = LL_dat[1,3], label = "0.001", hjust = 0, size = 8) +
  geom_segment(x = LL_dat[1,2], xend = LL_dat[1,2], y = 0, yend = LL_dat[1,3], linetype = "dashed")
```

## Likelihood for a sample

* Each subject has a likelihood for their observation
  * How to combine them?
* Probability theory: Joint probability of independent events is the **product** of the individual probabilities
  * Same works for likelihoods: 
  
$$L = \prod_{i=1}^{n} L_i = L_1 \times L_2 \times L_3 \times \cdots \times L_n$$

## What does likelihood mean?

* Individual
  * *Relatively* higher $L_i$ means that **that individual's score** is **more likely** than **other scores** (given the distributional assumptions)
  
* Sample
  * *Relatively* higher $L$ means that **the sample's scores** are **more likely** than **other samples of scores** (given the distributional assumptions)

## Sample likelihood

* Sample likelihood for this **very small** sample

$$L = 0.001 \times 0.003 \times 0.067 \times 0.078 \times 0.030 \times 0.067 \times 0.001$$

$$\times 0.030 \times 0.067 \times 0.030 = 0.000000000000000001327$$

* Especially as $n$ increases: Very very very small number, rounding errors
  * Computers have problems with this

## Log-likelihood

* Logarithms: $log(ab) = log(a) + log(b)$

$$L = \prod_{i=1}^{n} L_i = L_1 \times L_2 \times L_3 \times \cdots \times L_n$$
becomes

$$log(L) = \sum_{i=1}^{n} log(L_i) = log(L_1) + log(L_2) + log(L_3) + \cdots + log(L_n)$$

## Log-likelihood

```{r}
#| code-fold: true
round(LL_dat, 3)
```

* $L = 0.000000000000000001327$
* $log(L) = -7.028 + -5.908 + -2.708 + \cdots + -3.508 = -41.16$

## What does log-likelihood mean?

* Log-likelihood is **always negative**

* Individual
  * *Relatively* **higher** $log(L_i)$ means that **that individual's score** is **more likely** than **other scores** (given the distributional assumptions)
  
* Sample
  * *Relatively* **higher** $log(L)$ means that **the sample's scores** are **more likely** than **other samples of scores** (given the distributional assumptions)

## Log-likelihoods

```{r}
#| code-fold: true
ggplot(data = LL_dat, aes(x = BDI, y = LogLikelihood)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm",formula = y~poly(x,3), se = FALSE, color = "black")
```

## Distribution actually unknown!!!

* **We don't actually know the population values**
  * We draw samples and calculate statistics to **estimate** them
* **Conceptually**, with *maximum likelihood*, we try out different values of the mean and SD to figure out which are **most likely**
  * Which values result in the **highest (log)likelihood**

## Assume SD = 5

* Calculate the sample log-likelihood for several potential mean values
```{r}
#| code-fold: true
mu <- c(10:20)
logL <- c(-42.764,	-40.804,	-39.244,	-38.084,	-37.324,	-36.964,	-37.004,	-37.444,	-38.284,	-39.524,	-41.164)
aud_LL <- data.frame(mu, logL)
aud_LL
```

## Tangents and minimums again

* **Conceptually**, we try out different values of the mean and SD to figure out which are **most likely**
* **Actually**, we again use **calculus** and **derivatives**
  * Find the **maximum** of the likelihood function
  
## LL function maximized at 15.4

```{r}
#| code-fold: true
ggplot(data = aud_LL, aes(x = mu, y = logL)) +
  geom_smooth(method = "lm",formula = y~poly(x,3), se = FALSE, color = "black") + 
  labs(x = "Estimate of mean", y = "log(L)") +
  geom_hline(yintercept = -36.932, color = "dodgerblue", linetype = "dashed", linewidth = 1) +
  annotate("pointrange", x = 15.4, y = -36.932, ymin = -36.932, ymax = -36.932, color = "dodgerblue", size = 1)
```

## Extensions

* **More than 1 parameter** estimated
  * 1 parameter = 2D plots
  * k parameters = (k+1)D plots
* Can't assume the outcome is from a **normally distriubuted population**  
  * Replace **normal likelihood** (i.e., PDF) with appropriate version
  * A **slight** simplification

## Extensions

* A variation on maximum likelihood called **full information maximum likelihood** or "FIML" can be used to improve estimation in the presence of **missing data**
  * See the "References" section of the course website

## Caveats

* For linear regression, the **likelihood function** is very well-behaved
  * Shaped like upside down "U"
  * Only one maximum value
* For more complex models, the **likelihood function** is less well-behaved
  * Missing data
  * More complex, multivariate models
  * Non-normal outcomes

# Hypothesis testing

## Uncertainty

* **Sampling distributions** capture uncertainty in estimate
  * Sample is not population
  * Estimates vary from sample to sample
  * Uncertainty in the point estimate
* Produce quantities like $p$-values, confidence intervals
  * Based on the **standard error**
  * Standard error = standard deviation of sampling distribution
  
## Standard error estimation

* Use the **likelihood function**
  * **Second derivative** of the likelihood function
  * Tells you have quickly the *1st derivative* **changes**
  * Basically, how steep is the maximum likelihood **peak**

## Small SE vs large SE

:::: {.columns}

::: {.column width="50%"}

```{r}
#| code-fold: true
mu <- c(0:40)
L_smallse <- dnorm(mu, 20, 2)
L_largese <- dnorm(mu, 20, 5)
LL_smallse <- log(L_smallse)
LL_largese <- log(L_largese)
LLse_dat <- data.frame(mu, L_smallse, L_largese, LL_smallse, LL_largese)
#round(LLse_dat, 3)
ggplot(data = LLse_dat, aes(x = mu, y = LL_smallse)) +
  geom_smooth(method = "lm",formula = y~poly(x,3), se = FALSE, color = "black") +
  ylim(-10,0) +
  xlim(0,40) +
  annotate("pointrange", x = 15, y = -4.737, ymin = -4.737, ymax = -4.737, color = "dodgerblue", size = 1) +
  geom_segment(x = 15, xend = 15, y = -10, yend = -4.737, linetype = "dotted", linewidth = 1) +
  geom_hline(yintercept = -1.6) +
  labs(x = "Estimate of mean", y = "Log-likelihood")
```

:::

::: {.column width="50%"}

```{r}
#| code-fold: true
ggplot(data = LLse_dat, aes(x = mu, y = LL_largese)) +
  geom_smooth(method = "lm",formula = y~poly(x,3), se = FALSE, color = "black") +
  ylim(-10,0) +
  xlim(0,40) +
  annotate("pointrange", x = 15, y = -3.028, ymin = -3.028, ymax = -3.028, color = "dodgerblue", size = 1) +
  geom_segment(x = 15, xend = 15, y = -10, yend = -3.028, linetype = "dotted", linewidth = 1) +
  geom_hline(yintercept = -2.5) +
  labs(x = "Estimate of mean", y = "Log-likelihood")
```

:::

::::

## Information matrix

* a.k.a. Fisher Information Matrix
* $k \times k$ matrix, where $k$ is the number of parameters estimated
* Values on **diagonal** are **variances** of **parameter estimates**
  * *Square root* to get the **standard error**
* Error about "information matrix" = problem estimating standard errors

## Types of significance tests

* Equivalent in very large samples
  * **Wald tests**
  * **Likelihood ratio tests**

## Wald tests

* Usually used to test a **single parameter**
  * e.g., regression coefficient
* Divide estimate by its **standard error**
  * Compare to *appropriate hypothesized distribution*
* Regression coefficients use $t$-test, but may see $z$-tests
  * ML estimates are asymptotically normally distributed

## Likelihood ratio tests

* Used to test **1 or more** parameters
  * e.g., sets of predictors
  * Similar to `anova()` that we've used previously
* Compare likelihoods of two different **nested** models
  * Difference in -2*LL between models $\sim \chi^2(df)$
    * Degrees of freedom = difference in parameters
  * `anova(m1, m2, test = "LRT")`

## Degrees of freedom

* Observations are independent pieces of **information**
* Each thing we **estimate** *removes* something
  * One sample $t$-test: Degrees of freedom = $n - 1 = 10 - 1 = 9$
  * Regression coefficient: Degrees of freedom = $n - k - 1$
  * $F$-test of $R^2$: $k$, $n - k - 1$

## Example data
  
* `ICU` data from the **Stat2Data** package

```{r}
library(Stat2Data)
data(ICU)
head(ICU)
```

## Model 1

* `Age` and `Sex` predict `Pulse`

```{r}
m1 <- lm(data = ICU, Pulse ~ Age + Sex)
summary(m1)
```

## Degrees of freedom

* Regression coefficient
  * $b_{Age} = 0.04533, t(197) = 0.474, p = .636$
  * $n - k - 1$ (sample size minus predictors minus 1)
* Overall $F$-test 
  * $R^2 = 0.0027, F(2, 197) = 0.2667, p = .7662$
  * Degree of freedom 1: $k$ (number of predictors)
  * Degree of freedom 2: $n - k - 1$ (sample size minus predictors minus 1)

## Information matrix

* `vcov()` function in **stats** package (always loaded)
  * Values on main diagonal are **variances**
  * Square root to get standard error

```{r}
vcov(m1)
```

* $\sqrt{15.53330671}$ = 3.94123, which is standard error for `Sex`

## Model 2

* `Age` and `Sex` predict `Pulse`, then add `Infection` and `Emergency`

```{r}
m2 <- lm(data = ICU, Pulse ~ Age + Sex + Infection + Emergency)
summary(m2)
```

## Information matrix

```{r}
vcov(m2)
```

## Likelihood ratio test

* `logLik()` function from **stats** package (always loaded)

```{r}
logLik(m1)
logLik(m2)
```

* Why different degrees of freedom here?
  * Counts estimating the **variance** of the outcome as well
  * Intercept + number of predictors + variance

## Likelihood ratio test

```{r}
anova(m1, m2, test = "LRT")
-2 * (logLik(m1) - logLik(m2))
```

# Next time

## Interactions

* Non-additive effects
  * Relationship differs across groups
  * Easy to add to the model
    * Interpretation is more difficult

