---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Assess** models for **(multi)collinearity** among predictors
* **Conduct** outlier analyses to determine extreme and/or problematic cases

# Collinearity

## Collinearity

* Sometimes called *multicollinearity*
* **High correlations** between **predictors** in linear regression
* Why does that matter?

## Recall: Partial correlations

Correlation between $X_1$ and $Y$, no effect of $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (3,3) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\begin{scope}
\clip (0,0) circle (2cm);
\clip (3,0) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (3,3) circle (2cm);
\end{tikzpicture}
```

## Recall: Partial correlations

Correlation between $X_1$ and $Y$, removing the effect of $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (1.5,2.5) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\begin{scope}
\clip (0,0) circle (2cm);
\clip (3,0) circle (2cm);
\fill[white] (1.5,2.5) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (1.5,2.5) circle (2cm);
\end{tikzpicture}
```

## Recall: Partial correlations

Correlation between $X_1$ and $Y$, removing the effect of $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (1,2) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\begin{scope}
\clip (0,0) circle (2cm);
\clip (3,0) circle (2cm);
\fill[white] (1,2) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (1,2) circle (2cm);
\end{tikzpicture}
```


## Why "collinear"?

:::: {.columns}

::: {.column width="50%"}
* Variables are **vectors**
* Represent any vector as a **line** from the **origin** (0,0) to a **point**
* Angles between vectors are **correlations**
  * 180 degrees = -1
  * 90 degrees = 0
  * On top of one another = 1

:::

::: {.column width="50%"}
```{r, fig.width = 6}
#| echo: FALSE
#library(ggh4x)
dat2 <- data.frame(x=runif(10),y=runif(10))

tick_frame <- 
  data.frame(ticks = seq(-10, 10, length.out = 21), 
             zero=0) %>%
  subset(ticks != 0)

lab_frame <- data.frame(lab = seq(-10, 10),
                        zero = 0) %>%
  subset(lab != 0)

tick_sz <- (tail(lab_frame$lab, 1) -  lab_frame$lab[1]) / 128

ggplot(data = dat2, aes(x = x, y = y)) +
  # point 1
  geom_point(aes(x = 4, y = 3), color = "red", size = 4) +
  annotate("text", x = 4, y = 3.5, label = "(4,3)", size = 6) +
  geom_segment(aes(x = 0, y = 0, xend = 4, yend = 3),
                  arrow = arrow()) +
  # point 2
  geom_point(aes(x = -1, y = 5), color = "blue", size = 4) +
  annotate("text", x = -1.75, y = 5, label = "(-1,5)", size = 6) +
  geom_segment(aes(x = 0, y = 0, xend = -1, yend = 5),
                  arrow = arrow()) +
  # point 3
  geom_point(aes(x = 2, y = -2), color = "green", size = 4) +
  annotate("text", x = 2.5, y = -2.5, label = "(2,-2)", size = 6) +
  geom_segment(aes(x = 0, y = 0, xend = 2, yend = -2),
                  arrow = arrow()) +
  # point 4
  geom_point(aes(x = 1, y = -2), color = "purple", size = 4) +
  annotate("text", x = 1, y = -2.5, label = "(1,-2)", size = 6) +
  geom_segment(aes(x = 0, y = 0, xend = 1, yend = -2),
                  arrow = arrow()) +  
  xlim(c(-5, 5)) + ylim(c(-5, 5)) +
  theme(axis.line = element_line()) + 
  geom_segment(x = 0, xend = 0, 
               y = -10, yend = 10,
               size = 0.5) +
  geom_segment(y = 0, yend = 0, 
               x = -10, xend = 10,
               size = 0.5) +
  # point 5
  geom_point(aes(x = 2, y = 1.5), color = "cyan", size = 4) +
  annotate("text", x = 1.5, y = 2, label = "(2,1.5)", size = 6) +
  geom_segment(aes(x = 0, y = 0, xend = 2, yend = 1.5),
                  arrow = arrow()) +
  # x ticks
  geom_segment(data = tick_frame, 
               aes(x = ticks, xend = ticks, 
                   y = zero, yend = zero + tick_sz)) +
  # y ticks
  geom_segment(data = tick_frame, 
               aes(x = zero, xend = zero + tick_sz, 
                   y = ticks, yend = ticks)) + 
  # labels
  geom_text(data=lab_frame, aes(x=lab, y=zero, label=lab), vjust=1.5) +
  geom_text(data=lab_frame, aes(x=zero, y=lab, label=lab), hjust=1.5) +
  theme_void()
  #coord_axes_inside(labels_inside = TRUE)
```
:::

::::

## Data

* `ICU` data from the **Stat2Data** package

```{r}
library(Stat2Data)
data(ICU)
colnames(ICU)
```

## Correlations between variables

```{r}
cor(ICU)
```

## Predict `SysBP` from "everthing" 

```{r}
m1 <- lm(data = ICU, SysBP ~ . - ID - AgeGroup)
summary(m1)
```

## `Emergency` predicts `SysBP` alone

```{r}
m2 <- lm(data = ICU, SysBP ~ Emergency)
summary(m2)
```

## Variance inflation factor (VIF)

* VIF = $\frac{1}{1 - R^2_j}$
  * $R^2_j$ is $R^2$ for **all other predictors** predicting that predictor
* Standard error of a regression coefficient = $\frac{s^2}{(n - 1) \hat{var}(X_j)}  \times \frac{1}{1 - R^2_j}$
  * $s^2$ is the variance of the residuals
  * $n$ is the sample size
  * $\hat{var}(X_j)$ is the estimated variance of the predictor

## Variance inflation factor (VIF)

* Cut-offs for **high VIF**: 5 (common) or 10
* VIF is not the ratio of SEs for big model vs small model
  * Square root of VIF is how much larger the standard error is *compared to a model where the predictor has* **0 correlation** *to all other predictors*
* Might also / instead hear about **tolerance**
  * $\frac{1}{VIF}$ or $1 - R^2_j$

## Variance inflation factor (VIF)

```{r}
library(car)
vif(m1)
```

## Big model vs small model

* Two things change
  * Big model has **partial regression coefficients**
    * "Holding all other variables constant **(at 0)**"
    * Changes the **regression coefficient**: -8.16 vs -13.71
  * Variance inflated in big model due to **correlated predictors**
    * Changes the **standard error**: 5.59 vs 5.20
    * About 7.5% increase

## Highly correlated predictors

* `Age` and `AgeGroup` are correlated 0.899

```{r}
m3 <- lm(data = ICU, SysBP ~ . - ID)
library(car)
vif(m3)
```

## Summary: Collinearity

* Look at correlations between predictors
  * Not the whole story
  * VIF uses $R^2$ for **all other predictors** predicting that predictor
* Look at VIF values for the model
  * Cut off of 5
* Very high VIF
  * Do you **need all** of those predictors?
  * Can you **combine** some predictors?

# Outliers

## Outliers

* Observations that are 
  * *Extreme* on **predictor(s)**
  * *Extreme* on **predicted values**
  * *Change* **coefficients** or **predicted values**

## Outlier statistics

* **Model-based**
  * Based on a **linear regression** (or other) model
* **Case-wise**
  * **One** value per case (observation)
* Three types of outlier statistics for linear regression
  * Leverage
  * Discrepancy
  * Influence

## Leverage

* Extreme in the **predictor space**
  * "Hat matrix" from the model
* How far is a case from the **centroid** of the predictor space?
* Cut offs with $k$ predictors
  * Large samples, many predictors: $\frac{2(k + 1)}{n}$
  * Smaller samples: $\frac{3(k + 1)}{n}$

## Centroid of predictor space

* If just `Age` and `Pulse` are predictors

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Pulse, y = Age)) +
  geom_point(alpha = 0.3, 
             color = "blue",
             size = 4) +
  annotate("pointrange",
           x = mean(ICU$Pulse, na.rm = TRUE),
           y = mean(ICU$Age, na.rm = TRUE),
           ymin = mean(ICU$Age, na.rm = TRUE),
           ymax = mean(ICU$Age, na.rm = TRUE),
           color = "red",
           size = 2) +
  stat_ellipse(linetype = "dashed")
```

## Leverage

* `hatvalues()`
* `.hat` value: `augment()` from **broom** package

## Leverage: `hatvalues()`

```{r}
hatvalues(m1)
```

## Leverage: `augment()`

```{r}
library(broom)
m1_aug <- augment(m1)
m1_aug
```

## Histogram of leverage values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = .hat)) +
  geom_histogram()
```

## Index plot of leverage values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .hat)) +
  geom_point()
```

## Index plot of leverage values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .hat)) +
  geom_point() +
  geom_hline(yintercept = 3*(6+1)/200, 
             color = "red",
             linetype = "dashed") +
  geom_hline(yintercept = 2*(6+1)/200,
             color = "blue") +
  geom_text(aes(label=ifelse((.hat > 2*(6+1)/200), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```

## Discrepancy

* Extreme **predicted values**
  * Discrepancy between **observed** and **predicted** values
  * Residual = Observed - Predicted = $Y - \hat{Y}$
  * *Externally studentized residuals*
* How far is a case from the **regression line**?
* ES residual follows $t$ distribution with $n - k - 1$ df
  * Cut off: $\pm3$ for large sample, otherwise $\pm2$

## Raw residuals

* Residual = $e_i = Y_i - \hat{Y}_i$

* **Problems**
  * Metric depends on the **scale of the outcome**
  * An outlier **hides** by pulling the regression line toward itself

## Hiding outlier

```{r}
#| code-fold: true
x <- c(1:15)
y <- 2 * x + rnorm(15, 0, 8)
dat <- data.frame(x, y)
dat_out <- dat
dat_out[14,2] <- 50
dat_del <- dat_out
dat_del[14,2] <- NA

ggplot(data = dat_out,
       aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(data = dat_out, method = "lm", se = FALSE, color = "black") +
  ylim(0, 60)
```

## Hiding outlier

```{r}
#| code-fold: true

ggplot(data = dat_out,
       aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(data = dat_out, method = "lm", se = FALSE, color = "black") +
  geom_smooth(data = dat_del, method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  annotate("rect", xmin = 13.5, xmax = 14.5, ymin = 46, ymax = 54, fill = "red", alpha = 0.2) +
  ylim(0, 60)
```

## Internally studentized residual

* "Studentized" = standardized
  * Divide raw residual by its standard deviation
  * $SD_{e_i} = \sqrt{MS_{error}(1 - h_{ii})}$
* Internally studentized residual = $\frac{e_i}{SD_{e_i}}$

* **Problems**
  * **What distribution** does this follow???
  * Outlier pulls line toward itself, **increasing** $MS_{error}$

## Externally studentized residual

* **Deleted residual**: Difference between *observed value* and *predicted value* **in a regression in which the point was deleted**
* Deleted residual = $d_i = Y_i - \hat{Y}_{i(i)}$
  * Standard deviation based on deleted: $SD_{d} = \sqrt{MS_{error(i)}(1 - h_{ii})}$
* Externally studentized residual = $\frac{d_i}{SD_{d}}$

## Discrepancy: Externally studentized residuals

* `rstudent()` function from the **car** package

```{r}
library(car)
m1_stresid <- ICU %>% mutate(esresid = rstudent(m1))
head(m1_stresid)
```

## Histogram of ES residuals

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_stresid,
       aes(x = esresid)) +
  geom_histogram()
```

## Index plot of ES residuals

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_stresid,
       aes(x = c(1:nrow(m1_stresid)), 
           y = esresid)) +
  geom_point()
```

## Index plot of ES residuals

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_stresid,
       aes(x = c(1:nrow(m1_stresid)), 
           y = esresid)) +
  geom_point() +
  geom_hline(yintercept = 2, 
             color = "blue",
             linetype = "dashed") +
  geom_hline(yintercept = -2,
             color = "blue",
             linetype = "dashed") +
  geom_text(aes(label=ifelse((esresid > 2), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2) +
  geom_text(aes(label=ifelse((esresid < -2), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```

## What's up with `ID` = 921?

```{r}
library(kableExtra)
m1_aug %>% filter(ID == 921) %>%
  kable()
```

## Influence

* Changes **coefficients** and/or **predicted values**
  * *Predicted values*: **Cook's D** and DFFITS
  * *Regression coefficients*: DFBETAS
* Based on **case deletion**
* How much does the case **change** the **coefficient(s)** and/or the **predicted values**?
* Cutoff: Cook's D > 1

## Influence = leverage x discrepancy

```{r}
#| code-fold: true
ggplot(data = dat,
       aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") 
```

## Histogram of influence values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = .cooksd)) +
  geom_histogram()
```

## Index plot of Influence values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .cooksd)) +
  geom_point()
```

## Index plot of influence values

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .cooksd)) +
  geom_point() +
  geom_hline(yintercept = 1, 
             color = "red",
             linetype = "dashed") +
  geom_text(aes(label=ifelse((.cooksd > 1), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```

## Summary: Handling outliers

* Leverage and discrepancy will flag **most extreme ~5%**
* *Don't* base decisions on a single measure
  * Look for a **pattern** of suspicious cases
  * **Same case** shows up with high leverage, high discrepancy, high influence
* *Don't* just delete cases
  * *Sensitivity analysis*: **Same conclusions**?

## Repeat offenders?

:::: {.columns}

::: {.column width="33%"}

Leverage

```{r}
#| code-fold: true
m1_aug %>% filter(.hat > 2*(6+1)/200) %>% select(ID)
```

:::

::: {.column width="33%"}

Discrepancy

```{r}
#| code-fold: true
m1_stresid %>% filter(esresid > 2) %>% select(ID)
m1_stresid %>% filter(esresid < -2) %>% select(ID)
```

:::

::: {.column width="33%"}

Influence

```{r}
#| code-fold: true
m1_aug %>% filter(.cooksd > 1) %>% select(ID)
```

:::

::::

# Next time

## Some details

* Hypothesis testing
  * Sampling distribution, degrees of freedom
* Estimation of the model
  * $R^2$, likelihood
* Interaction (moderator) effects

