---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
library(broom)
library(ggdist) # logistic regression plot
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Conduct** logistic regression for *binary outcomes*
* **Interpret** results in terms of *probability*, *odds*, and *log-odds*

# Linear regression review

## Assumptions of linear regression

General linear model (GLM, linear regression, ANOVA) makes **three assumptions** about the **residuals** ($e_i = Y_i - \hat{Y}_i$) of the model

1.    **Independence**: observations (i.e., residuals) from different subjects **do not depend on one another**
2.    **Constant variance** (homoscedasticity): **variance of residuals is same** at all values of predictor(s)
3.    **Conditional normality**: **residuals are normally distributed** at each value of predictor(s)

## ICU data

```{r}
library(Stat2Data)
data(ICU)
head(ICU)
```

## Linear regression on normal outcome

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x = Age, 
           y = Pulse)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm")
```

## Assumptions met!

```{r}
#| code-fold: true
m1_norm <- lm(Pulse ~ Age, 
              data = ICU)
m1_norm_resid <- augment(m1_norm)
ggplot(data = m1_norm_resid, 
       aes(x = .resid)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 color = "blue") +
  geom_density()
```

## Assumptions met!

```{r}
#| code-fold: true
ggplot(data = m1_norm_resid, 
       aes(sample = .resid)) + 
  geom_qq() + 
  geom_qq_line() +
  labs(x = "Theoretical", 
       y = "Sample")
```

## Assumptions met!

```{r}
#| code-fold: true
ggplot(data = m1_norm_resid, 
       aes(x = Age, 
           y = .resid)) + 
  geom_point() + 
  geom_smooth()
```

# Binary variable

## A binary variable is not normal

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x = as.factor(Survive))) + 
  geom_bar()
```

## Plot of data with fit line

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x = Age, 
           y = Survive)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", 
              se = FALSE) +
  coord_cartesian(ylim = c(0,1)) 
```

## Plot of data with fit line

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x = Age, 
           y = Survive)) + 
  geom_jitter(height = 0.02, 
              alpha = 0.5) + 
  geom_smooth(method = "lm", 
              se = FALSE) +
  coord_cartesian(ylim = c(0,1)) 
```

<!-- ## Plot of data with fit line -->

<!-- ```{r} -->
<!-- #| fig-alt: "Binary outcome vs predictor with linear fit" -->
<!-- #| output-location: column -->
<!-- #| results: hold -->
<!-- #| echo: true -->
<!-- library(ggdist) -->
<!-- ggplot(data = ICU,  -->
<!--        aes(x = Age,  -->
<!--            y = Survive)) +  -->
<!--   stat_dots(data = ICU,  -->
<!--             aes(y = Survive,  -->
<!--                 side = ifelse(Survive == 0,  -->
<!--                               "top",  -->
<!--                               "bottom")),  -->
<!--             scale = 1/5) + -->
<!--   geom_smooth(method = "lm",  -->
<!--               se = FALSE) + -->
<!--   coord_cartesian(ylim = c(0,1))  -->
<!-- ``` -->

## Plot of residuals

```{r}
#| code-fold: true
m1_bin <- lm(Survive ~ Age, 
             data = ICU)
resids_m1_bin <- augment(m1_bin)
ggplot(data = resids_m1_bin, 
       aes(x = .resid)) + 
  geom_histogram(aes(y = ..density..), 
                 fill = "white", 
                 color = "blue") +
  geom_density() 

```

## Plot of residuals

```{r}
#| code-fold: true
ggplot(data = resids_m1_bin, 
       aes(sample = .resid)) + 
  geom_qq() + 
  geom_qq_line() +
  labs(x = "Theoretical", 
       y = "Sample")
```

## Plot of residuals

```{r}
#| code-fold: true
ggplot(data = resids_m1_bin, 
       aes(x = Age, 
           y = .resid)) + 
  geom_point() + 
  geom_smooth()
```

## What **NOT** to do

-   **Ignore** the problem

    -   Do linear regression anyway
    -   Call it **linear probability model**

-   **Transform** the *outcome*

    -   Square root, natural log, etc.
    -   May *slightly* normalize *univariate* residual distribution
    -   **Does not fix heteroscedasticity, (conditional) non-normality**

## A binary variable is not normal

```{r}
#| fig-alt: "Distribution of binary variable" 
#| echo: false
ggplot(data = ICU, 
       aes(x = as.factor(Survive))) + 
  geom_bar()
```

## What to do

* **Generalized linear model (GLiM)**
  * Not a single model but a **family** of regression models
  * **Choose** features (e.g., residual distribution) to match the **characteristics** of your outcome variable
  * Accommodates many **continuous** and **categorical** outcome variables
  * Includes **logistic regression** and **Poisson regression**
  * Transformation of the **predicted** outcome

# Logistic regression

## (Binary) logistic regression

* **Outcome**: binary 
  * Observed value ($Y$): 0 or 1, where 1 = "success" or "event"
  * Predicted value ($\hat{Y}$): **Probability** of success, between 0 and 1
*   **Residual/outcome distribution**: binomial
*   **Link function**: logit (or log-odds) = $ln\Big(\frac{\hat{Y}}{1 - \hat{Y}}\Big) = ln\left(\frac{\hat{p}}{1-\hat{p}}\right)$

$$ln\left(\frac{\hat{p}}{1-\hat{p}}\right) = b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p$$

## Reminder: normal distribution

$$f(x) = {\frac {1}{{\sqrt {2\pi \sigma^2}}}}e^{-{\frac {(x-\mu)^2 }{2 \sigma^2 }}}$$

Mean of normal distribution = $\mu$

Variance of normal distribution = $\sigma^2$

-   Mean and variance are **different parameters** and are **unrelated**
  * **Homoscedasticity**: Constant variance doesn't depend on mean

<!-- ## Binomial distribution -->

<!-- $$P(X = k) = {n \choose k} p^k (1-p)^{n-k}$$ -->

<!-- -   $n$ is the sample size -->
<!-- -   $p$ is the probability of an event\ -->
<!-- -   $k$ is the observed number of events\ -->
<!-- -   ${n \choose k} = \frac{n!}{k!(n-k!)}$ and is read as "$n$ choose $k$" -->

## Binomial distribution

Probability of $k$ events in $n$ trials, each with probability $p$ of event?

::: columns
::: column
* $p$ = 0.5, $n$ = 10

```{r}
#| code-fold: true
n <- 10
p <- 0.5
k <- seq(0, 10, 1)
Prob <- dbinom(k, n, p)
bin0_5 <- data.frame(k, Prob)
ggplot(data = bin0_5, aes(x = k, y = Prob)) + 
  geom_col() +
  scale_x_continuous(breaks=seq(0,10,1)) +
  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5)) +
  ylim(0,0.4) +
  labs(y = "P(X = k)")
```
:::

::: column
* $p$ = 0.1, $n$ = 10

```{r}
#| code-fold: true
n <- 10
p <- 0.1
k <- seq(0, 10, 1)
Prob <- dbinom(k, n, p)
bin0_2 <- data.frame(k, Prob)
ggplot(data = bin0_2, aes(x = k, y = Prob)) + 
  geom_col() +
  scale_x_continuous(breaks=seq(0,10,1)) +
  scale_y_continuous(breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5)) +
  ylim(0,0.4) + 
  labs(y = "P(X = k)")
```
:::
:::

## Binomial distribution

$$P(X = k) = {n \choose k} p^k (1-p)^{n-k}$$

Mean of a binomial distribution: $np$\
Variance of a binomial distribution: $np(1-p)$

-   Mean and variance are **related** to one another

    -   They are functions of the same parameters ($n$ and $p$)
    
- **Heteroscedasticity is built into logistic regression**

## Logistic regression: What we model

* Linear regression: Model the **mean** of the outcome
  * Observed and predicted outcome are the *same type of variable*
* Logistic regression: Model the **probability of a "success" or "event"**
  * Observed is **binary** (**0 or 1**) 
  * Predicted is **probability** (**from 0 to 1**)
  * From the **probability**, we can also get the **odds** of a success and the **logit or log-odds** of a success

## Logistic regression model

* `glm()` function in **stats** package (always loaded)
  * `data`: same as `lm()`
  * `formula`: same as `lm()`
  * `family`: Residual distribution and link function

```{r}
ICU <- ICU %>% mutate(AgeC = Age - mean(Age, na.rm = TRUE))
m1 <- glm(data = ICU, 
          Survive ~ AgeC, 
          family = binomial(link = "logit"))
```

```{r}
#| code-fold: true
m1_b0 <- round(coefficients(m1)[1], 3)
m1_b1 <- round(coefficients(m1)[2], 3)
```

## Logistic regression model

:::: {.columns}

::: {.column width="30%"}
* Estimate
* Standard error
* **$z$ statistic**
* $p$-value
* **No $R^2$**
* Deviance value
:::

::: {.column width="70%"}
```{r}
summary(m1)
```
:::

::::





<!-- ## Figure: What we model -->

<!-- ```{r} -->
<!-- #| code-fold: true -->
<!-- m1 <- glm(Survive ~ Age, family = binomial(link = "logit"), data = ICU) -->
<!-- m1_b0 <- round(coefficients(m1)[1], 3) -->
<!-- m1_b1 <- round(coefficients(m1)[2], 3) -->
<!-- #m1_b0 -->
<!-- #m1_b1 -->
<!-- ICU$pred <- (exp(m1_b0 + m1_b1 * ICU$Age)) / (1 + exp(m1_b0 + m1_b1 * ICU$Age)) -->
<!-- ICU$low <- ICU$pred - ICU$pred * (1-ICU$pred) -->
<!-- ICU$high <- ICU$pred + ICU$pred * (1-ICU$pred) -->
<!-- ggplot(ICU, aes(x=Age, y=Survive)) + -->
<!--   geom_point() + -->
<!--   stat_smooth(method="glm", method.args=list(family="binomial"), se=FALSE) + -->
<!--   geom_ribbon(aes(ymin=low, ymax=high), alpha = 0.3) -->
<!-- ``` -->

## Figure: What we model

```{r}
#| code-fold: true
ggplot(ICU, 
       aes(x=AgeC, 
           y=Survive)) +
  geom_point() +
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE)
```

## Three forms of logistic regression

::: {.columns}

::: {.column width="20%"}
Probability: 
:::

::: {.column width="80%"}
$$\hat{p} = \frac{e^{(\color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p})}}{1+e^{(\color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p})}}$$ 
:::

:::

::: {.columns}

::: {.column width="20%"}
Odds:
:::

::: {.column width="80%"}
$$\hat{odds} = \frac{\hat{p}}{1-\hat{p}} = e^{\color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p}}$$ 
:::

:::

::: {.columns}

::: {.column width="20%"}
Logit:
:::

::: {.column width="80%"}
$$ln\left(\frac{\hat{p}}{1-\hat{p}}\right) = \color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p}$$
:::

:::

# Probability metric

## What is probability ($p$)?

- Likelihood of a "success" or "event"
- Ranges from 0 to 1 (inclusive)
- Both options are equally likely when $p = 0.5$

## $\hat{p} = \frac{e^{`r m1_b0` + `r m1_b1` X}}{1 + e^{`r m1_b0` + `r m1_b1` X}}$

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(ICU, 
       aes(x=AgeC, 
           y=Survive)) + 
  geom_point() + 
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE) 
```

## $\hat{p} = \frac{e^{`r m1_b0` + `r m1_b1` X}}{1 + e^{`r m1_b0` + `r m1_b1` X}}$

```{r}
#| output-location: column
#| results: hold
#| echo: true
library(ggdist)
ggplot(ICU, 
       aes(x=AgeC, 
           y=Survive)) + 
  #geom_point() + 
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE) +
  stat_dots(data = ICU, 
            aes(y = Survive, 
                side = ifelse(Survive == 0, 
                              "top", 
                              "bottom")), 
            scale = 1/5)
```

## Interpreting probability

* **Intercept**: $b_0$ is related to the **probability of success** when **X = 0**
  * $b_0$ \> 0: **Success (1) more likely** than failure (0) when X = 0
  * $b_0$ \< 0: **Failure (0) more likely** than success (1) when X = 0
* **Slope**: $b_1$ tells you how **predictor X relates to probability of success**
  * $b_1$ \> 0: Probability of a **success increases as X increases**
  * $b_1$ \< 0: Probability of a **success decreases as X increases**

## Interpreting probability: Example

$$\hat{p} = \frac{e^{\color{OrangeRed}{`r m1_b0`} + `r m1_b1` X}}{1 + e^{\color{OrangeRed}{`r m1_b0`} + `r m1_b1` X}}$$

* Interpretation of example **intercept**
  * $b_0$ > 0: **Success (1) more likely than failure (0) when X = 0**
  * Probability of success when X = 0: $\frac{e^{\color{OrangeRed}{b_0}}}{1 + e^{\color{OrangeRed}{b_0}}} = \frac{e^{\color{OrangeRed}{`r m1_b0`}}}{1 + e^{\color{OrangeRed}{`r m1_b0`}}} =`r round(eval(exp(m1_b0)) / (1 + exp(m1_b0)), 3)`$
  * Probability of survival at mean age = `r round(eval(exp(m1_b0)) / (1 + exp(m1_b0)), 3)`

## P(success|X=0)

```{r}
#| code-fold: true
int_eval <- round(eval(exp(m1_b0)) / (1 + exp(m1_b0)), 3)

ggplot(ICU, 
       aes(x=AgeC, 
           y=Survive)) + 
  geom_point() + 
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE) +
  annotate("segment", 
           x = -2, xend = 0, 
           y = int_eval, yend = int_eval, 
           colour = "black", linetype = "dashed") +
  annotate("segment", 
           x = 0, xend = 0, 
           y = 0, yend = int_eval, 
           colour = "black", linetype = "dashed") +
  annotate("pointrange", 
           x = 0, y = int_eval, 
           ymin = int_eval, ymax = int_eval, 
           colour = "black", size = 1.5) +
  annotate("text", 
           x = 0, y = int_eval + 0.1, 
           colour = "black", label = int_eval, size = 6)
```

## Interpreting probability: Example

$$\hat{p} = \frac{e^{`r m1_b0` + \color{OrangeRed}{`r m1_b1`} X}}{1 + e^{`r m1_b0` + \color{OrangeRed}{`r m1_b1`} X}}$$

* Interpretation of example **slope**
  * $b_1$ < 0: **Probability of a success decreases as X increases**
  * Probability of survival decreases as age increases
    * **But** it's non-linear

## Probability metric is non-linear

* Linear regression:
  * **Constant**, linear slope
  * Slope depends on the **slope only**
* Logistic regression (probability):
  * **Non-linear** slope
  * No interaction but **conditional effect**

<!-- ## Probability metric is non-linear -->

```{r, echo = FALSE}
pred_prob_1.5 <- signif(exp(m1_b0 + m1_b1 * 1.5) / (1 + exp(m1_b0 + m1_b1 * 1.5)), 3)
```

<!-- When $\color{blue}{X = 1.5}$: -->

<!-- $$\hat{P}(success) = \hat{p} = \frac{e^{b_0 + b_1 \color{blue}{X}}}{1+e^{b_0 + b_1 \color{blue}{X}}} = \frac{e^{`r m1_b0` + `r m1_b1` \times \color{blue}{1.5}}}{1 + e^{`r m1_b0` + `r m1_b1` \times \color{blue}{1.5}}} = `r signif(eval(exp(m1_b0 + m1_b1 * 1.5) / (1 + exp(m1_b0 + m1_b1 * 1.5))), 3)`$$ -->

<!-- Approximate **slope** at that point is -->

<!-- $$\hat{p} (1-\hat{p}) \color{OrangeRed}{b_1} = `r pred_prob_1.5` \times (1 - `r pred_prob_1.5`) \times \color{OrangeRed}{`r m1_b1`} = `r signif(eval(pred_prob_1.5 * (1 - pred_prob_1.5) * m1_b1), 3)`$$ -->

## Probability metric is non-linear

```{r, echo = FALSE}
x_val <- c(-30, -20, -10, 0, 10, 20, 30)
pred_prob_slope <- exp(m1_b0 + m1_b1 * x_val) / (1 + exp(m1_b0 + m1_b1 * x_val))
#slope <- pred_prob_slope * (1 - pred_prob_slope) * m1_b1
slopes <- data.frame(x_val, pred_prob_slope)
colnames(slopes) <- c("AgeC", "Predicted probability")
rownames(slopes) <- NULL

knitr::kable(slopes, digits = 4, row.names = F)
```

## Probability metric is non-linear

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x=AgeC, 
           y=Survive)) + 
  geom_point() + 
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE)
```

## Probability metric is non-linear

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x=AgeC, 
           y=Survive)) + 
  geom_point() + 
  stat_smooth(method="glm", 
              method.args=list(family="binomial"), 
              se=FALSE, 
              fullrange = T) +
  xlim(-100, 250)
```

## A caution about probability equation 

:::{.callout-warning}
You might also see the probability defined as $\hat{p} = \frac{1}{1 + e^{-({b_{0} + b_{1} X})}}$

Or more generally, $\hat{p} = \frac{1}{1 + e^{-(Xb)}}$

* These are **numerically equivalent** to what we've talked about

  * But did you notice the negative sign?
  * No? You didn't expect it and missed it in the complicated equation?
  * Yeah, that's why we don't use this version
:::

# Odds metric

## What are odds?

* *Odds* is the **ratio** of **two probabilities**

$$odds = \frac{\hat{p}}{(1 - \hat{p})}$$

* Odds is the ratio of probability of a "success" ($\hat{p}$) to the probability of "not a success" $(1 − \hat{p})$ 

## How do odds work?

* Probability ranges from 0 to 1, switches at 0.5

  * **Success more likely** than failure when $p > 0.5$
  * **Success less likely** than failure when $p < 0.5$

* Odds range from $0$ to $+\infty$, switches at 1 

  * **Success more likely** than failure when $odds > 1$
  * **Success less likely** than failure when $odds < 1$

## $\hat{odds} = \frac{\hat{p}}{(1 - \hat{p})} = e^{`r m1_b0` + `r m1_b1` X}$

```{r}
#| code-fold: true
ICU <- ICU %>% mutate(pred_odds = exp(m1_b0 + m1_b1 * AgeC),
                      pred_p = exp(m1_b0 + m1_b1 * AgeC) / (1 + exp(m1_b0 + m1_b1 * AgeC)),
                      obs_odds = pred_p / (1 - pred_p))

ggplot(data = ICU, 
       aes(x = AgeC, 
           y = obs_odds)) +  
  geom_line(color="blue", 
            aes(x = AgeC, 
                y = pred_odds)) +
  labs (y = "Odds") +
  annotate("segment", 
           x = -40, xend = 35, 
           y = 1, yend = 1, 
           colour = "black", linetype = "dashed")
```

## Interpreting odds

* **Intercept**: $b_0$ is related to the **odds of success when $X$ = 0**   
  * Odds of success **when X = 0**: $e^{b_0}$
  * $b_0$ > 0: Odds of success > 1 when $X$ = 0  
  * $b_0$ < 0: Odds of success < 1 when $X$ = 0  
* **Slope**: $b_1$ tells you how predictor X relates to **odds of success**
  * $b_1$ > 0: Odds of success **increases** as $X$ **increases** 
  * $b_1$ < 0: Odds of a success **decreases** as $X$ **increases** 

## Interpreting odds: Example

$$\hat{odds} = \frac{\hat{p}}{(1 - \hat{p})} = e^{\color{OrangeRed}{`r m1_b0`} + `r m1_b1` X}$$

* Interpretation of example **intercept**
  * $b_0 > 0$ so odds of success > 1 when $X$ = 0  
  * Success (1) more likely than failure (0) when $X$ = 0   
  * Odds of success when $X$ = 0: $e^{\color{OrangeRed}{b_0}} = e^{\color{OrangeRed}{`r m1_b0`}} = `r signif(eval(exp(m1_b0)), 3)`$   
    * A "success" is about `r signif(eval(exp(m1_b0)), 3)` times as likely as a "failure"
    * Compare to `r signif(eval(int_eval), 3)` **probability** of success: `r signif(eval(int_eval), 3)` / `r signif(eval(1 - int_eval), 3)` = `r signif(eval(int_eval / (1 - int_eval)), 3)`

## Interpreting odds: Example

$$\hat{odds} = \frac{\hat{p}}{(1 - \hat{p})} = e^{`r m1_b0` + \color{OrangeRed}{`r m1_b1`} X}$$

* Interpretation of example **slope**
  * $b_1$ < 0: Odds of a success **decreases** as $X$ **increases**
  * **Odds** of survival decreases as age increases
    * But it’s non-linear

## Odds metric is nonlinear

* This non-linear change is presented in terms of **odds ratio**
  * Constant, **multiplicative** change in predicted odds
  * For a 1-unit difference in $X$, the **predicted odds** of success is **multiplied by** the **odds ratio**
* *Example*: odds ratio $= e^{b_1}= e^{`r m1_b1`} = `r signif(eval(exp(m1_b1)), 3)`$
  * For a 1-unit difference in $X$, the **predicted odds** of success is **multiplied by** $`r signif(eval(exp(m1_b1)), 3)`$
  * Each 1-unit increase in $X$ **reduces** the odds of a success
  
## Odds ratio = `r signif(eval(exp(m1_b1)), 3)`)  

```{r}
#| code-fold: true
ggplot(data = ICU, 
       aes(x = AgeC, 
           y = obs_odds)) +  
  geom_line(color="blue", 
            aes(x = AgeC, 
                y = pred_odds)) +
  labs (y = "Odds") +
  annotate("segment", 
           x = -40, xend = 35, 
           y = 1, yend = 1, 
           colour = "black", linetype = "dashed")
```

## Odds metric is nonlinear

```{r}
#| echo: false
slopes <- slopes %>% mutate(odds_slope = pred_prob_slope / (1 - pred_prob_slope))
colnames(slopes) <- c("AgeC", "Predicted probability", "Predicted odds")

knitr::kable(slopes, digits = 4, row.names = F)
```

## A caution about odds 

:::{.callout-warning}
* **Odds ratios** are very popular in medicine and epidemiology
* They can be **extremely** misleading
* The **same odds ratio** corresponds to many **different** probability values
  * Odds ratio $= \frac{odds = 3}{odds = 1} = 3$
    * Corresponds to probability of 0.75 vs 0.5
  * Odds ratio $= \frac{odds = 9}{odds = 3} = 3$  
    * Corresponds to probability of 0.90 vs 0.75
:::

# Logit or log-odds metric

## What is the logit?

* **Logit** or **log-odds** is the natural log ($ln$) of the odds
  * As **probability** of "success" increases (nonlinearly, S-shaped curve)
  * The *odds* of "success" increases (also nonlinearly, exponentially)
  * The **logit** of "success" increases **linearly**
  
## How does the logit work?

* Probability ranges from 0 to 1, **switches at 0.5**
* Odds range from 0 to $+\infty$ , **switches at 1**
* Logit ranges from $-\infty$ to $+\infty$, **switches at 0**
  * Success more likely than failure when logit > 0
  * Success less likely than failure when logit < 0

## $ln\left(\frac{\hat{p}}{(1 - \hat{p})}\right) = `r m1_b0` + `r m1_b1` X$

```{r}
#| code-fold: true
ggplot(ICU, aes(x=AgeC, y=obs_odds)) +  
  geom_line(color="blue", aes(x = AgeC, y = log(pred_odds))) +
  labs (y = "Logit") +
  annotate("segment", x = -40, xend = 35, y = 0, yend = 0, colour = "black", linetype = "dashed")
```

## Interpreting logit

* **Intercept**: $b_0$ is related to the **logit of success** when **X = 0**
  * Logit of success when X = 0: $b_0$
    * $b_0$ > 0: Logit > 0 when X = 0
    * $b_0$ < 0: Logit < 0 when X = 0
* **Slope** : $b_1$ is the relationship between **predictor X** and **logit of success**  
  * $b_1$ > 0: Logit of a success **increases** as X **increases**
  * $b_1$ < 0: Logit of a success **decreases** as X **increases**

## Interpreting logit: Example

$$\hat{logit} = ln\left(\frac{\hat{p}}{(1 - \hat{p})}\right) = \color{OrangeRed}{`r m1_b0`} + `r m1_b1` X$$

* Interpretation of example **intercept**
  * $b_0$ > 0: Logit > 0 when X = 0
  * Logit of success when X = 0: $\color{OrangeRed}{b_0} = \color{OrangeRed}{`r m1_b0`}$
  * Success more likely than failure when X = 0

## Interpreting logit: Example

$$\hat{logit} = ln\left(\frac{\hat{p}}{(1 - \hat{p})}\right) = `r m1_b0` + \color{OrangeRed}{`r m1_b1`} X$$

* Interpretation of example **slope**
  * $b_1$ < 0: Logit of a success **decreases** by $\color{OrangeRed}{`r m1_b1`}$ units when X increases by 1 unit

# Miscellaneous

## So which metric should I use?

* They are **equivalent** and give you the same information
  * Just packaged differently
* Use the metric that
  * Makes the most *sense* to you
  * You can *explain* fully
  * Is most commonly used in your *field*

## Metrics: Some things to keep in mind

* Odds ratios tell you about **change**, but not **where you start**

  * If you report odds ratios, *also report some measure of probability* e.g., probability of success at the mean of X
  * 10x change is $5%$ to $50%$ or $0.05%$ to $0.5%$?
  
* Logit is nice because it’s **linear**, but it’s not very **interpretable** 

  * What is a "logit"? It’s just a *mathematical* concept that makes a straight line -- not actually **meaningful**
  * *But many measures don’t have meaningful metrics...*

## Why a $z$-statistic?

* Maximum likelihood estimation
  * Estimates are assumed to be normally distributed

* Known variance
  * Binomial distribution: mean = $np$, variance = $np(1-p)$
  * Use $t$ distribution when you need to **estimate** variance
    * Use $z$ distribution when you don't

## Predictors in logistic regression

* Similar to linear regression
  * Continuous: Center for interpretability
  * Categorical: Code appropriately
  * Coefficients are **partial** coefficients
* Tests are $z$-tests, not $t$-tests

## Models in logistic regression

* Add predictors in sets
  * Likelihood ratio tests to compare models
* Coefficients are in **logit metric**
  * So are the **confidence intervals**
  * Confidenct intervals in other metrics
* $R^2_{multiple}$ analogue
* Packages to make some of these things easier

## Interactions in logistic regression

* Effects (in probability metric) are already nonlinear   
  * Already conditional
* Interactions (also conditional) are complicated here
  * Conditional on top of conditional
  * Can have interactions *without* a product term
  * Can have no interaction *with* a significant product term
* See McCabe et al. (2022): Full reference on the References page

# Summary

## Summary

* Use logistic regression when your outcome is binary

  * **Don't use linear regression**

* Be careful with interpretation no matter what
  * **Probability**: Probability *makes sense*, but it's *nonlinear*
  * **Odds**: Odds ratio *seems to make sense* but it can be *misleading*
  * **Logit**: *Linear* but what even is a *logit*?

* But many basic concepts parallel linear regression
  * Intercept, slope(s), adding predictors, $R^2_{multiple}$

## In class

* Run some logistic regression models
* Interpret the results
* Examples of the things we didn't get to
  * $R^2_{multiple}$, compare models, report full results

## Next time

* ~~Logistic regression for **binary** outcomes~~
* Poisson regression for **count** outcomes
* Survival analysis for **survival** outcomes

