---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* Describe the logic of **linear regression** 
* Describe the **assumptions** of linear regression
* Briefly **present results** of linear regression
* Use a **regression equation** to summarize a model
* Compare and contrast **observed**, **predicted**, and **residual** values

# Motivation

## Linear regression

* General linear model or **GLM**
  * Includes linear regression, ANOVA, $t$-tests
* One **dependent variable** or **outcome**
  * Predicted by 1 or more continuous or categorical **independent variables** or **predictors**
* Causality not assumed, but sometimes (wrongly) implied
  
## Purposes of linear regression

* **Inference**
  * Confirm effect of predictor(s) on outcome
* **Prediction**
  * Predict outcome value based on predictor value(s)
* **Variance explained**
  * How much of the variance in the outcome is explained by the predictor(s)?

## Assumptions

* **One** *continuous* outcome variable
* **One or more** predictor variable (nominal, ordinal, interval, ratio)
* Relationship is **linear in the predictors**
* Observations are **independent** (uncorrelated)
* **Residuals** are *conditionally normally distributed* with *constant variance* (homoscedasticity)

# Example

## Data

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

## Data

```{r}
library(Stat2Data)
data(ICU)
head(ICU)
```

## Relations between variables

* Y versus X
* Y as a function of X
* `Y ~ X`
* X predicts Y

## Example 1: `Pulse` vs `Age`

* Does `Pulse` depend on `Age`?

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Age, y = Pulse)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm",
              se = TRUE)
```

## Example 2: `Pulse` vs `Emergency`

* Do emergency and non-emergency admits have similar `Pulse` rates?

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Emergency, y = Pulse)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm", 
              se = TRUE)
```

# Inference

## `lm()` function  

* `lm()` function in **stats** package (loaded by default)
* Inputs
  * `data`: data frame
  * `formula`: relationship to model (`Y ~ X`)
* Outputs
  * `coefficients`, `residuals`, etc.
  * See [documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm) 

## Example 1: `Pulse` vs `Age`

```{r}
pulse_vs_age <- lm(data = ICU, Pulse ~ Age)
summary(pulse_vs_age)
```

## Example 1: `Pulse` vs `Age`

* Intercept = Expected `Pulse` value when `Age` = 0
  * $b_0$ = 96.048, $t(1)$ = 16.594, $p$ < .0001
  * `Pulse` for a person at `Age` = 0 is 96.048
* Slope = Expected change in `Pulse` for a 1 unit difference in `Age`
  * $b_1$ = 0.049, $t(1)$ = 0.526, $p$ > .05
  * For each 1 unit increase in `Age`, we expect `Pulse` to increase by 0.526 points (NS)

## Example 2: `Pulse` vs `Emergency`

```{r}
pulse_vs_ER <- lm(data = ICU, Pulse ~ Emergency)
summary(pulse_vs_ER)
```

## Example 2: `Pulse` vs `Emergency`

* Intercept = Expected `Pulse` when `Emergency` = 0
  * $b_0$ = 91.113, $t(1)$ = 25.049, $p$ < .0001
  * `Pulse` for a person w non-emergency admit is 91.113
* Slope = Expected change in `Pulse` for a 1 unit difference in `Emergency`
  * $b_1$ = 10.628, $t(1)$ = 2.505, $p$ < .05
  * For each 1 unit increase in `Emergency`, we expect `Pulse` to increase by 10.628 points
  * **Group difference** (0 to 1) is 10.628 points

## Regression equations

* Two forms
  * $\hat{Y} = b_0 + b_1 X_1 + b_2 X_2 + \cdots$
    * **Predicted Y** is a function of coefficients and predictors
  * $Y = b_0 + b_1 X_1 + b_2 X_2 + \cdots + error$
    * **Observed Y** is a function of coefficients and predictors, **plus error** 

## Regression equations

* Example 1: 
  * $\hat{Pulse} = 96.048 + 0.049*Age$
* Example 2: 
  * $\hat{Pulse} = 91.113 + 10.628*Emergency$

## How to describe slopes

* For each **1 unit increase** (or **1 unit change** or **1 unit difference**) in `X`, we expect `Y` to increase by $b_1$ points on average
  * Are we **increasing** or **changing** `X`? 
  * In these models (and often), no
* For **two individuals who are 1 unit apart** on `X`, we expect their scores on `Y` to differ by $b_1$ points on average
  * This is a *mouthful* but more technically accurate

## Reporting $p$-values

* Non-significant effect
  * Thresholds: $b_1$ = 0.049, $t(1)$ = 0.526, $p$ > .05
  * Exact $p$-value: $b_1$ = 0.049, $t(1)$ = 0.526, $p$ = .599
  * Not significant: $b_1$ = 0.049, $t(1)$ = 0.526, NS
* Significant effect
  * Thresholds: $b_1$ = 10.628, $t(1)$ = 2.505, $p$ < .05
  * Exact $p$-value: $b_1$ = 10.628, $t(1)$ = 2.505, $p$ = .0131
* $p$-value is **never** 0: Report $p$ < .001 or similar

# Prediction

## Observed and predicted values

* **Observed** values of the outcome in the dataset
  * One value per observation
  * Observed outcome: $Y$
* **Predicted** values of the outcome that are **predicted** by the model
  * One value per observation
  * Predicted outcome: $\hat{Y}$
  * Use the regression equation to calculate

## Predicted values

* Example 1: `Pulse ~ Age`
  * `Age` = 87: $\hat{Pulse} = 96.048 + 0.049 * 87 = 100.311$
  * `Age` = 27: $\hat{Pulse} = 96.048 + 0.049 * 27 = 97.371$
  * `Age` = 59: $\hat{Pulse} = 96.048 + 0.049 * 59 = 98.939$
  * `Age` = 0: $\hat{Pulse} = 96.048 + 0.049 * 0 = 96.048$
  * Etc. for each value of `Age`

## Predicted values

* Example 2: `Pulse ~ Emergency`
  * Emergency = 0: $\hat{Pulse} = 91.113 + 10.628 * 0 = 91.113$
  * Emergency = 1: $\hat{Pulse} = 91.113 + 10.628 * 1 = 101.741$

## Predicted values

* `predict()` function in R
  * Talk more about this in coming weeks
  * Can use to create predicted values for
    * Original data
    * All possible values
    * Specific values

## Example 1: Observed and predicted

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Age, y = Pulse)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm",
              se = FALSE, 
              size = 2)
```

## Example 2: Observed and predicted

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Emergency, y = Pulse)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              size = 2)
```
  
## Residual values

* Residuals are the *difference* between the **observed** and **predicted** values **for an observation**
  * One residual per observation
  * Residual: $Y - \hat{Y}$
* Assumptions for **residuals**
  * *Constant variance* (homoscedasticity)
  * *Conditionally normally distributed*

## Predicted and residual values

```{r}
ICU$pred1 <- fitted(pulse_vs_age)
ICU$resi1 <- residuals(pulse_vs_age)
ICU$pred2 <- fitted(pulse_vs_ER)
ICU$resi2 <- residuals(pulse_vs_ER)
head(ICU)
```

## What is the residual?

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Age, y = Pulse)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm",
              se = FALSE, 
              size = 2) +
  geom_segment(aes(x = Age,
                   xend = Age,
                   y = Pulse,
                   yend = pred1))
```

## Constant variance?

* Plot residual vs predictor

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Age, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Constant variance?

* Plot residual vs predicted (for more than 1 predictor)

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = pred1, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Normally distributed?

* Q-Q plot (overall distribution)

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU,
       aes(sample = resi1)) + 
       stat_qq() +
       stat_qq_line()
```

## Normally distributed?

* Histogram

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU,
       aes(x = resi1)) +
  geom_histogram()
```

## (Conditionally) Normally distributed?

* Histogram **at each level of predictor**

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU,
       aes(x = resi2)) +
  geom_histogram() +
  facet_grid(cols = vars(Emergency))
```

## Constant variance?

* Plot residual vs predictor

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Emergency, y = resi2)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

# Variance explained

## R-squared multiple

* One number summary of the predictive power of the model
  * Proportion: Ranges from 0 to 1
    * 0 = predictor(s) explain **no variance** in outcome
    * 1 = predictor(s) explain **all variance** in outcome
* **Squared correlation** (between observed and predicted values)\
  * **Effect size** for the *model*
  * Other definitions related to *estimation*

## Correlations

:::: {.columns}

::: {.column width="50%"}

* Correlation = 0.2 
  * $R^2$ = 0.04

```{r}
#| echo: false
library(faux)
cmat <- c(1, .2, 0, 0,
          .2, 1, 0, 0,
          0, 0, 1, .8,
          0, 0, .8, 1)
dat <- rnorm_multi(n = 100, 
                  mu = c(0, 0, 0, 0),
                  sd = c(1, 1, 1, 1),
                  r = cmat, 
                  varnames = c("A", "B", "C", "D"),
                  empirical = TRUE)
ggplot(data = dat, 
       aes(x = A, y = B)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

::: {.column width="50%"}

* Correlation = 0.8
  * $R^2$ = 0.64

```{r}
#| echo: false
ggplot(data = dat, 
       aes(x = C, y = D)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

::::

## R-squared multiple

* We'll talk more about $R^2_{multiple}$
  * Estimation
  * Effect size
  * Logistic regression and Poisson regression


# Presenting results

## Packages to show model results

* `summary()` function from **base** R
* `summ()` function from **jtools** package
* `modelsummary()` function in **modelsummary** package
  * Nice HTML output, multiple models in one table
* Non-exhaustive list
  
## Example 1: `summary()`

```{r}
summary(pulse_vs_age)
```

## Example 1: `summ()`

```{r}
library(jtools)
summ(pulse_vs_age)
```

## Example 1: `modelsummary()`

```{r}
library(modelsummary)
modelsummary(pulse_vs_age)
```

## Example 1: `summary()` + `kableExtra()`

```{r}
library(kableExtra)
kable(summary(pulse_vs_age)$coefficients)
```

## Example 2: `summary()`

```{r}
summary(pulse_vs_ER)
```

## Example 2: `summ()`

```{r}
library(jtools)
summ(pulse_vs_ER)
```

## Example 2: `modelsummary()`

```{r}
library(modelsummary)
modelsummary(pulse_vs_ER)
```

## Example 2: `summary()` + `kableExtra()`

```{r}
library(kableExtra)
kable(summary(pulse_vs_ER)$coefficients)
```

## Multiple models in `modelsummary()`

```{r}
#| echo: false
pulse_age_ER <- lm(data = ICU, Pulse ~ Age + Emergency)
```


```{r}
library(modelsummary)
modelsummary(list(pulse_vs_age, pulse_vs_ER, pulse_age_ER))
```

## Example 1: Full results

* The intercept ($b_0$) was 96.048, $t(1)$ = 16.594, $p$ < .0001. An individual who was 0 years old is expected to have a pulse rate of 96.048. 
* The slope for age ($b_1$) was 0.049, $t(1)$ = 0.526, $p$ > .05. For each 1 year increase in age, the expected pulse rate increased by 0.048 points; this effect was not significant.

* The $R^2$ for this model was 0.0014, F(1, 198) = 0.2969, p > .05. Less than 1% of the variance in pulse rate was explained by age; this was not significant.

## Example 2: Full results

* The intercept ($b_0$) was 91.113, $t(1)$ = 25.049, $p$ < .0001. An individual who had a non-emergency admission is expected to have a pulse rate of 91.113. 

* The slope for type of admission ($b_1$) was 10.628, $t(1)$ = 2.505, $p$ < .05. The difference in expected pulse rate between non-emergency admissions (0) and emergency admissions (1) was 10.628 points. 

* The $R^2$ for this model was 0.0307, F(1, 198) = 6.275, p < .05. About 3% of the variance in pulse rate was explained by admission type.

# Next time

## Expanding the model

* More complicated models
  * Two or more predictors
  * Non-linear relationships, interaction (moderator) effects
* Predictor coding and centering
* Comparing models
* More interpretation, more checking assumptions
* Estimation of the model

