---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Run models** with interaction effects to assess conditional effects of predictors
* **Probe** interaction models to fully understand all effects

# Additive models

## Our models so far

* Additive effects
  * The effect of *two variables* is the **sum** of *two individual effects*, nothing more
  * Effect of predictor 1: Add that effect
  * Effect of predictor 2: Add that effect
  * Effect of **both** predictors: Add **both** effects

## Example data and analysis

```{r}
library(Stat2Data)
data(ICU)
m1 <- lm(data = ICU, 
         Pulse ~ Emergency + Age)
library(broom)
ICU_aug <- augment(m1)
```

## Additive effects

```{r}
summary(m1)
```

## Constant, additive effects

$$\hat{Pulse} = 84.913 + 11.452 Emergency + 0.097 Age$$

* Effect of `Emergency` is the same regardless of value of `Age`
* Effect of `Age` is the same regardless of value of `Emergency`
  * Same **slope** wrt `Age` regardless of value of `Emergency`
  * `Emergency` = 0: $\hat{Pulse} = 84.913 + 0.097 Age$
  * `Emergency` = 1: $\hat{Pulse} = 96.365 + 0.097 Age$

## Additive effects = parallel lines

```{r}
#| code-fold: true
ggplot(data = ICU_aug,
       aes(x = Age, y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.4, size = 2) +
  geom_line(aes(y = .fitted, color = as.factor(Emergency)), 
            linewidth = 1) +
  labs(color = "Emergency")
```

## Assumptions: Residuals

```{r}
#| code-fold: true
ggplot(data = ICU_aug,
       aes(x = .fitted, 
           y = .std.resid)) +
  geom_point() +
  geom_smooth()
```

## Assumptions: Residuals

```{r}
#| code-fold: true
ggplot(data = ICU_aug,
       aes(x = .fitted, 
           y = .std.resid, 
           color = as.factor(Emergency))) +
  geom_point() +
  geom_smooth() +
  labs(color = "Emergency")
```

## More going on?

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.4, 
             size = 2) +
  geom_smooth(se = FALSE,
              aes(color = as.factor(Emergency))) +
  labs(color = "Emergency")
```

## More going on?

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.4, 
             size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              aes(color = as.factor(Emergency))) +
  labs(color = "Emergency")
```

# Interactions (moderation)

## Interaction is a product term

$$\hat{Pulse} = b_0 + b_1 Emergency + b_2 Age + b_3 Emergency\times Age$$

* `Emergency` and `Age` have their own effects
* But they do something special **together**
* The effect of *one predictor* is **conditional on** (or **depends on**) the value of the *other predictor*
* Non-parallel lines

## Interaction effect

```{r}
m2 <- lm(data = ICU, Pulse ~ Emergency + Age + Emergency*Age)
summary(m2)
```

## Interaction effect

$$\hat{Pulse} = 125.609 + -36.051 Emergency +$$
$$-0.541 Age + 0.761 Emergency\times Age$$

* Simplify for the two `Emergency` groups
  * Substitute 0 and 1 for `Emergency`

## Interaction effect

* `Emergency` = 0
  * $\hat{Pulse} = 125.609 + -36.051 (0) + -0.541 Age + 0.761 (0)\times Age$
  * $\hat{Pulse} = 125.609 + -0.541 Age$
* `Emergency` = 1
  * $\hat{Pulse} = 125.609 + -36.051 (1) + -0.541 Age + 0.761 (1)\times Age$
  * $\hat{Pulse} = (125.609 -36.051) + (-0.541 + 0.761) Age$
  * $\hat{Pulse} = 89.558 + 0.22 Age$
  
## What do all the coefficients mean?

$$\hat{Pulse} = 125.609 + -36.051 Emergency +$$
$$-0.541 Age + 0.761 Emergency\times Age$$

* Intercept: 125.609
  * Expected `Pulse` when all predictors are equal to 0
  * `Emergency` = 0 and `Age` = 0
  * `Pulse` for non-emergency admit newborn
  
## What do all the coefficients mean?

$$\hat{Pulse} = 125.609 + -36.051 Emergency +$$
$$-0.541 Age + 0.761 Emergency\times Age$$

* `Emergency` effect = -36.051
  * **Difference** in `Pulse` between emergency and non-emergency admits **when `Age` equals 0**
  * In this case, an *intercept* difference  
  
## What do all the coefficients mean?

$$\hat{Pulse} = 125.609 + -36.051 Emergency +$$
$$-0.541 Age + 0.761 Emergency\times Age$$
  
* `Age` effect = -0.541
  * Expected change in `Pulse` for a 1 unit change in `Age` **when `Emergency` equals 0**
  * *Slope* wrt `Age` for `Emergency` = 0 group

## What do all the coefficients mean?

$$\hat{Pulse} = 125.609 + -36.051 Emergency +$$
$$-0.541 Age + 0.761 Emergency\times Age$$
  
* Interaction = 0.761
  * **Difference** in `Age` slopes between emergency and non admits
  * Significant interaction = **slopes are different**
  * Not parallel lines
  
## Plot of interaction

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.3, 
             size = 2) +
  geom_abline(slope = -0.541, 
              intercept = 125.609, 
              linewidth = 1,
              color = "#F8766D") +
  geom_abline(slope = 0.22, 
              intercept = 89.558,
              linewidth = 1,
              color = "#00BA38") +
  xlim(0, 90) +
  theme(legend.position="none")
```

$\hat{Pulse} = 125.609 + -36.051 Emergency + -0.541 Age + 0.761 Emergency\times Age$
  
## Plot of interaction

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.3, 
             size = 2) +
  geom_abline(slope = -0.541, 
              intercept = 125.609, 
              linewidth = 1,
              color = "#F8766D") +
  geom_abline(slope = 0.22, 
              intercept = 89.558,
              linewidth = 1,
              color = "#00BA38") +
  xlim(0, 90) +
  theme(legend.position="none") +
  annotate("pointrange", 
           x = 0, y = 125.609, 
           ymin = 125.609, ymax = 125.609, 
           size = 1) +
  annotate("text", x = 0, y = 135, label = "125.609", size = 6)
```

$\hat{Pulse} = 125.609 + -36.051 Emergency + -0.541 Age + 0.761 Emergency\times Age$

## Plot of interaction

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.3, 
             size = 2) +
  geom_abline(slope = -0.541, 
              intercept = 125.609, 
              linewidth = 1,
              color = "#F8766D") +
  geom_abline(slope = 0.22, 
              intercept = 89.558,
              linewidth = 1,
              color = "#00BA38") +
  xlim(0, 90) +
  theme(legend.position="none") +
  annotate("pointrange", 
           x = 0, y = 125.609, 
           ymin = 125.609, ymax = 125.609, 
           size = 1) +
  annotate("text", x = 0, y = 135, label = "125.609", size = 6) +
  geom_segment(x = 0, xend = 0, 
               y = 89.558, yend = 125.609,
               linewidth = 1, linetype = "dashed") +
  annotate("text", x = 5, y = 105, label = "-36.051", size = 6)
```

$\hat{Pulse} = 125.609 + -36.051 Emergency + -0.541 Age + 0.761 Emergency\times Age$

## Plot of interaction

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.3, 
             size = 2) +
  geom_abline(slope = -0.541, 
              intercept = 125.609, 
              linewidth = 1,
              color = "#F8766D") +
  geom_abline(slope = 0.22, 
              intercept = 89.558,
              linewidth = 1,
              color = "#00BA38") +
  xlim(0, 90) +
  theme(legend.position="none") +
  annotate("pointrange", 
           x = 0, y = 125.609, 
           ymin = 125.609, ymax = 125.609, 
           size = 1) +
  annotate("text", x = 0, y = 135, label = "125.609", size = 6) +
  geom_segment(x = 0, xend = 0, 
               y = 89.558, yend = 125.609,
               linewidth = 1, linetype = "dashed") +
  annotate("text", x = 5, y = 105, label = "-36.051", size = 6) +
  geom_abline(intercept = 125.609, slope = -0.541,
              linetype = "dashed", linewidth = 1) +
  annotate("text", x = 25, y = 125, label = "-0.541", size = 6)
```

$\hat{Pulse} = 125.609 + -36.051 Emergency + -0.541 Age + 0.761 Emergency\times Age$

## Plot of interaction

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.3, 
             size = 2) +
  geom_abline(slope = -0.541, 
              intercept = 125.609, 
              linewidth = 1,
              color = "#F8766D") +
  geom_abline(slope = 0.22, 
              intercept = 89.558,
              linewidth = 1,
              color = "#00BA38") +
  xlim(0, 90) +
  theme(legend.position="none") +
  annotate("pointrange", 
           x = 0, y = 125.609, 
           ymin = 125.609, ymax = 125.609, 
           size = 1) +
  annotate("text", x = 0, y = 135, label = "125.609", size = 6) +
  geom_segment(x = 0, xend = 0, 
               y = 89.558, yend = 125.609,
               linewidth = 1, linetype = "dashed") +
  annotate("text", x = 5, y = 105, label = "-36.051", size = 6) +
  geom_abline(intercept = 125.609, slope = -0.541,
              linetype = "dashed", linewidth = 1) +
  annotate("text", x = 25, y = 125, label = "-0.541", size = 6) +
  geom_abline(intercept = 89.558, slope = 0.22,
              linetype = "dashed", linewidth = 1) +
  geom_curve(x = 80, xend = 80, 
             y = 82.329, yend = 107.158,
             linewidth = 1, linetype = "dashed",
             arrow = arrow(length = unit(0.3, "cm"), 
                           type = "closed")) +
  annotate("text", x = 87, y = 95, label = "0.761", size = 6)
```

$\hat{Pulse} = 125.609 + -36.051 Emergency + -0.541 Age + 0.761 Emergency\times Age$

## Center continuous predictors

* Recommended to **mean center** continuous predictors
  * Interpretable intercept
  * Interpretable lower order effects
  * Reduces collinearity between interaction and other predictors
* Use centered predictor in all places in the model
  
## Center continuous predictors

```{r}
ICU <- ICU %>% mutate(AgeC = Age - mean(Age, na.rm = TRUE))
m3 <- lm(data = ICU, Pulse ~ Emergency + AgeC + Emergency*AgeC)
summary(m3)
```

## What changes?

* Lower order terms because **"...when `Age` = 0"**
  * Intercept: 125.609 $\rightarrow$ 94.482
    * Now at mean of `Age` instead of `Age` = 0
  * Effect of `Emergency`: -36.051 $\rightarrow$ 7.754
    * Now at mean of `Age` instead of `Age` = 0
* Collinearity
  * Standard errors are **reduced** with centering for some terms

## Collinearity before and after centering

```{r}
library(car)
vif(m2)
vif(m3)
```

* Though don't worry about this **too** much, esp for product term
  * [https://statisticalhorizons.com/multicollinearity/](https://statisticalhorizons.com/multicollinearity/)
  * [https://search.r-project.org/CRAN/refmans/car/html/vif.html](https://search.r-project.org/CRAN/refmans/car/html/vif.html)
  
## What doesn't change?

* Higher order terms
  * Interaction: 0.761
  * Effect of `Age` (slope): -0.541
* $R^2_{multiple}$ and associated tests
* Log-likelihood

## Uncentered `Age`

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.4, 
             size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              aes(color = as.factor(Emergency))) +
  labs(color = "Emergency")
```

## Centered `Age`

```{r}
#| code-fold: true
ggplot(data = ICU,
       aes(x = AgeC, 
           y = Pulse, 
           group = Emergency)) +
  geom_point(aes(color = as.factor(Emergency)), 
             alpha = 0.4, 
             size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE,
              aes(color = as.factor(Emergency))) +
  labs(color = "Emergency")
```

## Probing the interaction

* **Simple slopes**: Slopes at specific values
  * `Emergency` = 0: $\hat{Pulse} = 125.609 + -0.541 Age$
  * `Emergency` = 1: $\hat{Pulse} = 89.558 + 0.22 Age$
* We know (from the original analysis)
  * Significance of coefficients for "group = 0" 
  * Significance of **differences** between the groups
* **What about "group = 1"?**

## Two approaches

* **Method 1**: *Recode your group variable* from 0,1 to 1,0
  * Now the group that **was** 1 is now 0
  * Interpret the intercept and slope for that group
* **Method 2**: *Calculate things*
  * Uses the variances and covariance of coefficients (`vcov()`)
  * See Aiken & West (1991), Preacher, Curran, Bauer (2006)

## Packages to make this easier for you

* Non-exhaustive list of packages
* **interactions**: [https://cran.r-project.org/package=interactions](https://cran.r-project.org/package=interactions)
* **marginaleffects**: [https://cran.r-project.org/package=marginaleffects](https://cran.r-project.org/package=marginaleffects)
  * Also [https://marginaleffects.com/](https://marginaleffects.com/)
* **sjPlot**: [https://cran.r-project.org/package=sjPlot](https://cran.r-project.org/package=sjPlot)

## Plot simple slopes

```{r}
#| output-location: column
#| results: hold
#| echo: true
library(interactions)
interact_plot(m3, 
              pred = AgeC, 
              modx = Emergency)
```

## Plot simple slopes with data

```{r}
#| output-location: column
#| results: hold
#| echo: true
interact_plot(m3, 
              pred = AgeC, 
              modx = Emergency, 
              plot.points = TRUE)
```

## Simple slopes analysis

```{r}
#| output-location: column
#| results: hold
#| echo: true
sim_slopes(m3, 
           pred = AgeC, 
           modx = Emergency, 
           johnson_neyman = FALSE)
```

## Johnson-Neyman intervals

* For **which values of `AgeC`** are the `Emergency` groups **different**?
  * Not about the **slopes** but about the **values**
* Johnson, P. O., & Fay, L. C. (1950). The Johnson-Neyman technique, its theory and application. *Psychometrika*, 15, 349–367.

## Johnson-Neyman intervals

```{r}
#| output-location: column
#| results: hold
#| echo: true
sim_slopes(m3, 
           pred = AgeC, 
           modx = Emergency, 
           johnson_neyman = TRUE)
```

## Johnson-Neyman intervals: Improved

```{r}
#| output-location: column
#| results: hold
#| echo: true
sim_slopes(m3, 
           pred = Emergency, 
           modx = AgeC, 
           johnson_neyman = TRUE)
```

## Johnson-Neyman intervals: Plot

```{r}
#| output-location: column
#| results: hold
#| echo: true
johnson_neyman(m3, 
               pred = Emergency, 
               modx = AgeC, 
               alpha = .05)
```

# Miscellaneous

## Main effects vs simple effects

* Factorial ANOVA: **Main effects** + interaction
  * **Main effect**: Effect of a predictor **averaged across** the levels of the other predictors
  * *Overall effect*
* Linear regression: **Simple (main) effects** + interaction
  * **Simple effect**: Effect of a predictor **at a specific value** of the other predictors
  * *Conditional effect*

## Main effects vs simple effects

* In general: Don't use the term "main effects" for regression
  * "Simple slopes" for the actual simple slopes
  * "Lower order effects" for the coefficients (i.e., $b_1$ and $b_2$)
* If interaction is significant
  * Lower order effects are conditional, so make sure to talk about them accordingly

## Two continuous predictors

```{r}
ICU <- ICU %>% mutate(SysBPC = SysBP - mean(SysBP, na.rm = TRUE))
m4 <- lm(data = ICU, Pulse ~ AgeC + SysBPC + AgeC*SysBPC)
summary(m4)
```

## Simple slopes

```{r}
#| output-location: column
#| results: hold
#| echo: true
interact_plot(m4, 
              pred = AgeC, 
              modx = SysBPC, 
              plot.points = TRUE)
```


# Next time

## Non-normal outcomes

* Logistic regression for **binary** outcomes
* Poisson regressoin for **count** outcomes
* Survival analysis for **survival** outcomes

