---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
library(broom)
library(ggdist) # logistic regression plot
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Conduct** Poisson regression for *count outcomes*
* **Interpret** results in terms of *predicted counts*
* **Describe** when to use alternative models / specifications
  * Negative binomial, inflated zeroes, offset, etc.

# Linear regression review

## Assumptions of linear regression

* Linear regression (GLM) makes 3 assumptions about the residuals ($e_i = Y_i - \hat{Y}_i$) of the model
  * Independence
  * Constant variance
  * Conditional normality

# Count variables

## What is a count?

* **Count variables** occur any time you **count** the number of times an event occurs in a *fixed period of time*
    * Number of visits to doctor in past 6 months
    * Number of days of poor QOL per month
    * Number of cigarettes smoked in a day
    * Number of gene repeats in a sample
    * Number of adverse events in a study

## Count outcomes in medical research

```{r}
#| code-fold: true
count1 <- rpois(n = 1000, lambda = 1.5)
count_dat <- as.data.frame(count1)
ggplot(data = count_dat, aes(count1)) +
    geom_bar(fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "Visits to doctor in past 6 months", y = "Freqency") +
    scale_x_continuous(breaks = c(0:12))
```

## Count variable properties

* Always
    * Discrete (whole numbers)
    * Lower bound of 0
* Typically
    * Low mean
    * High frequency of 0 values
    * Right skewed
    * Heteroskedasticity

## Example data: Alcohol use

* Simulated data[^a] made to resemble real data[^b]
* Outcome: **Number** of alcoholic drinks consumed *last Saturday*
  * Ranges from 0 to 17
* Predictors
    * Sensation seeking: 3 to 7 (theoretically 1 to 7)
    * Sex: 0 = female, 1 = male

[^a]: Simulated data in Coxe, West, & Aiken (2009)
[^b]: Real data in DeHart, Tennen, Armeli, Todd, & Affleck (2008)

## A count variable is not normal

```{r}
#| code-fold: true
library(readr)
dat <- read_table("jpapoisson.txt", 
    col_names = FALSE)
colnames(dat) <- c("ID", "sensation", "sex", "alcohol")
ggplot(data = dat, aes(x = alcohol)) +
    geom_bar(fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "Number of alcoholic drinks", y = "Frequency")
```

## Linear regression on count

```{r}
#| code-fold: true
ggplot(data = dat, aes(x = sensation, y = alcohol)) +
    geom_point(size = 3, 
               alpha = 0.4, 
               color = "#3697aa") +
  geom_smooth(method = "lm", 
              se = FALSE,
              color = "black") + 
  labs(x = "Sensation seeking", y = "Number of alcoholic drinks")
```

## Residuals for linear regression

```{r}
#| code-fold: true
m1_lin <- lm(data = dat, alcohol ~ sensation)
resids_m1_lin <- augment(m1_lin)
ggplot(data = resids_m1_lin, 
       aes(x = .resid)) + 
  geom_histogram(aes(y = ..density..), 
                 fill = "#3697aa", 
                 color = "#76777a") +
  geom_density() 
```

## Residuals for linear regression

```{r}
#| code-fold: true
ggplot(data = resids_m1_lin, 
       aes(sample = .resid)) + 
  geom_qq() + 
  geom_qq_line() +
  labs(x = "Theoretical", 
       y = "Sample")
```

## Residuals vs predictor

```{r}
#| code-fold: true
ggplot(data = resids_m1_lin, 
       aes(x = sensation, 
           y = .resid)) + 
  geom_point() + 
  geom_smooth()
```

## Residuals vs predictor

```{r}
#| code-fold: true
ggplot(data = resids_m1_lin, 
       aes(x = sensation, 
           y = .resid)) + 
  geom_point() + 
  geom_smooth() +   
  geom_abline(intercept = 0.75, 
                slope = -0.75, 
                color = "#dc1e34", 
                size = 1,
                linetype = "dashed") +
    geom_abline(intercept = 1.5, 
                slope = 1.9, 
                color = "#dc1e34", 
                size = 1,
                linetype = "dashed")
```

## What **NOT** to do

* **Ignore** the problem
  * Do linear regression anyway
  * Get biased estimates and out-of-range prediction

* **Transform** the *outcome*
  * Square root, natural log, etc.
  * May *slightly* normalize *univariate* residual distribution
  * **Does not fix heteroscedasticity, (conditional) non-normality**

## Transform the outcome - ugh!

```{r}
#| code-fold: true
dat2 <- dat %>%
    mutate(lnalc = log(alcohol + 0.01))
ggplot(data = dat2, aes(x = lnalc)) +
    geom_histogram(fill = "#3697aa", 
                   color = "#76777a", 
                   size = 1) +
    labs(x = "ln(Number of alcoholic drinks)", y = "Frequency")
```

## Transform the outcome - ugh!

```{r}
#| code-fold: true
dat2 <- dat2 %>%
    mutate(sqrtalc = sqrt(alcohol))
ggplot(data = dat2, aes(x = sqrtalc)) +
    geom_histogram(fill = "#3697aa", 
                   color = "#76777a", 
                   size = 1) +
    labs(x = "Square root(Number of alcoholic drinks)", y = "Frequency")
```

## What to do

* **Generalized linear model (GLiM)**
  * Not a single model but a **family** of regression models
  * **Choose** features (e.g., residual distribution) to match the **characteristics** of your outcome variable
  * Accommodates many **continuous** and **categorical** outcome variables
  * Includes **logistic regression** and **Poisson regression**
  * Transformation of the **predicted** outcome

# Poisson regression

## Poisson regression

* **Outcome**: count in a *fixed period of time*
  * Discrete, lower bound of zero, often right skewed
* **Residual / outcome distribution**: Poisson
* **Link function**: natural log = $ln(\hat{Y}) = ln(\hat{\mu})$

$$ln(\hat{\mu}) = b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p$$

## Reminder: normal distribution

$$f(x) = {\frac {1}{{\sqrt {2\pi \sigma^2}}}}e^{-{\frac {(x-\mu)^2 }{2 \sigma^2 }}}$$

Mean of normal distribution = $\mu$

Variance of normal distribution = $\sigma^2$

* Mean and variance are **different parameters** and are **unrelated**
  * **Homoscedasticity**: Constant variance doesn't depend on mean

## Poisson distribution

* Developed by Simeon Denis Poisson (1781 - 1840)
* **Number of times** a **low probability** event happens
  * Number of soldiers killed by horse-kicks each year in each corps in the Prussian cavalry (Bortkiewicz)
  * Number of yeast cells used when brewing Guinness beer
  * Number of typographical errors on each page of a book
  * Number of bombs falling in London in each block during Blitz
  * Number of earthquakes per day in Southern California

## Poisson distribution

$$P(X = k) = \frac{\lambda^k}{k!}e^{-\lambda}$$

* **Poisson distribution** is
  * *Discrete*
  * *Undefined for values less than 0*
  * *Right skewed* when *mean is low*
  * $\lambda =$ mean *and* variance: Heteroscedasticity

## Poisson: mean = variance = 1

```{r}
#| code-fold: true
pois_1 <- rpois(10000, 1)
pois_1_data <- data.frame(pois_1)
ggplot(data = pois_1_data, aes(x = pois_1)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)),
             fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "X", y = "Proportion") + 
    xlim(-1, 20) + 
    ylim(0, .400)
```

## Poisson: mean = variance = 4

```{r}
#| code-fold: true
pois_4 <- rpois(10000, 4)
pois_4_data <- data.frame(pois_4)
ggplot(data = pois_4_data, aes(x = pois_4)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)),
             fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "X", y = "Proportion") + 
    xlim(-1, 20) + 
    ylim(0, .400)
```

## Poisson: mean = variance = 7

```{r}
#| code-fold: true
pois_7 <- rpois(10000, 7)
pois_7_data <- data.frame(pois_7)
ggplot(data = pois_7_data, aes(x = pois_7)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)),
             fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "X", y = "Proportion") + 
    xlim(-1, 20) + 
    ylim(0, .400)
```

## Poisson: mean = variance = 10

```{r}
#| code-fold: true
pois_10 <- rpois(10000, 10)
pois_10_data <- data.frame(pois_10)
ggplot(data = pois_10_data, aes(x = pois_10)) + 
    geom_bar(aes(y = (..count..)/sum(..count..)), 
             fill = "#3697aa", color = "#76777a", size = 1) +
    labs(x = "X", y = "Proportion") + 
    xlim(-1, 20) + 
    ylim(0, .400)
```

## Poisson regression model

* `glm()` function in **stats** package (always loaded)
  * `data`: same as `lm()`
  * `formula`: same as `lm()`
  * `family`: Residual distribution and link function

```{r}
dat <- dat %>% mutate(sensationC = sensation - mean(sensation, na.rm = TRUE))
m1 <- glm(data = dat, 
          alcohol ~ sensationC, 
          family = poisson(link = "log"))
```

```{r}
#| code-fold: true
m1_b0 <- round(coefficients(m1)[1], 3)
m1_b1 <- round(coefficients(m1)[2], 3)
```

## Poisson regression model

:::: {.columns}

::: {.column width="30%"}
* Estimate
* Standard error
* **$z$ statistic**
* $p$-value
* **No $R^2$**
* Deviance
:::

::: {.column width="70%"}
```{r}
summary(m1)
```
:::

::::

## Two forms of Poisson regression

Predicted count:

$$\hat{\mu} = e^{\color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p}}$$ 

Natural log of predicted count:

$$ln(\hat{\mu}) = \color{OrangeRed}{b_0 + b_1 X_1 + b_2 X_2 + \cdots + b_p X_p}$$

## Predicted count interpretation

$$\hat{\mu} = e^{b_0 + b_1 X}$$

* Intercept
  * $e^{b_0}$ is the **predicted count** when **$X$ = 0**
* Slope
  * $e^{b_1}$ is the **rate ratio**
  * For a 1 unit increase in $X$, the predicted count is **multiplied by** $e^{b_1}$ 


## Example: $\hat{\mu} = e^{`r m1_b0` + `r m1_b1` \times sensationC}$

* Intercept = $e^{b_0} = e^{`r m1_b0`} = `r round(exp(m1_b0), 3)`$
  * People with mean value of `sensation` are expected to drink `r round(exp(m1_b0), 3)` alcoholic drinks
* Slope = $e^{b_1} = e^{`r m1_b1`} = `r round(exp(m1_b1), 3)`$
  * **Multiplicative** effect of $X$ on the predicted count
  * For a one unit increase in `sensation`, the number of predicted alcoholic drinks is **multiplied** by `r round(exp(m1_b1), 3)`
  * Also: `r (round(exp(m1_b1), 3) - 1)*100`$\%$ increase in the predicted outcome for each one unit increase in `sensation`


## Predicted counts: $e^{b_1} = `r round(exp(m1_b1), 3)`$

```{r}
#| code-fold: true
sens_vals <- data.frame(sensationC = c(-2, -1, 0, 1)) 
pred_counts <- predict(m1, newdata = sens_vals, type = "response")
sensationC <- c(-2, -1, 0, 1)
pred_alc <- c(1.82, 2.29, 2.89, 3.64)
pred_counts <- data.frame(sensationC, pred_alc)
ggplot(dat, 
       aes(x=sensationC, 
           y=alcohol)) +
  geom_point(color = "#3697aa",
             alpha = 0.3,
             size = 2) +
  stat_smooth(method="glm", 
              method.args=list(family="poisson"), 
              se=FALSE,
              color = "black") +
  geom_point(data = pred_counts, 
             aes(x = sensationC, y = pred_alc),
             size = 3,
             color = "black") +
  annotate("text", x = -2, y = 1.82, 
           label = "1.82", vjust = -1, size = 6) +
  annotate("text", x = -1, y = 2.29, 
           label = "2.29", vjust = -1, size = 6) +
  annotate("text", x = 0, y = 2.89, 
           label = "2.89", vjust = -1, size = 6) +
  annotate("text", x = 1, y = 3.64, 
           label = "3.64", vjust = -1, size = 6)
```

## Confidence intervals: Natural log

* Original coefficients and confidence intervals
  * Natural log of predicted count metric

```{r}
coef(m1)
confint(m1)
```

* Do they contain 0?

## Confidence intervals: Predicted count

* Exponentiate estimates and confidence intervals
  * **Predicted count** metric

```{r}
exp(coef(m1))
exp(confint(m1))
```

* Do they contain 1?

# (Over)Dispersion

## Equidispersion

* Poisson distribution displays **equidisersion**
  * Conditional mean = conditional variance = $\lambda$
* In practice in data, the actual variance of $Y$ is typically **larger** than expected (i.e., larger than the mean)
  * This is called **overdispersion**
  
## Why does overdispersion happen?

* Sometimes, just because
* Due to omitted predictors
  * Something we should include but are not
  * New predictors affect the **mean**
    * Heteroscedasticity: They can impact variance
* "Excess" zeroes
  * **Looks like** more variance 

## Why do we care?

* The variance estimated in the model ($\lambda$) helps determine the **standard errors** for regression coefficients
* It also plays a role in calculating **deviance**
* **If the variance in our data is larger than $\lambda$, the standard errors are too small and the deviance is wrong**
  * Test of significance, pseudo-$R^2$, model comparisons are **wrong**

## Is there overdispersion?

* Quick check: Residual deviance / degrees of freedom > 1

```{r}
summary(m1)
```

## Adding a predictor

* Can reduce overdispersion

```{r}
m1add <- glm(data = dat, 
          alcohol ~ sensationC + sex, 
          family = poisson(link = "log"))
summary(m1add)
```


## What do we do?

* There are two main alternative models for overdispersion
  * **Overdispersed Poisson regression** a.k.a. quasi-Poisson regression
    * *Same coefficients*, adjusted standard errors
  * **Negative binomial regression**
    * Different coefficients, different standard errors
* Both handle overdispersion, but in slightly different ways

## Overdispersed Poisson regression

* "Fudge factor"
  * Variance is larger than the model says, so multiply variance by something to make it larger
* Scale parameter: $\psi = \frac{\chi^2_{Pearson}}{residual\ df}$ or $\frac{deviance}{residual\ df}$
  * If equidispersion holds: $\psi = 1$
  * If overdispserion: $\psi > 1$
  * Compared to Poisson: Each standard error $\times \sqrt{\psi}$

## Overdispersed Poisson regression

```{r}
m2 <- glm(data = dat, 
          alcohol ~ sensationC, 
          family = quasipoisson)
summary(m2)
```


## Compare Poisson to OD Poisson

:::: {.columns}

::: {.column width="50%"}

* Poisson regression

```{r}
summary(m1)
```

:::

::: {.column width="50%"}

* Overdispersed Poisson

```{r}
summary(m2)
```

:::

::::

## Negative binomial regression

* Negative binomial distribution
  * Poisson distribution + gamma distribution
  * Adds **additional variance**
* Different scale parameter: $\alpha$, but in R is $\theta = 1/\alpha$
  * Mean = $\mu$
  * Variance = $\mu + \alpha \mu^2$
  * If overdispersion: $\alpha$ > 0
  * If not: $\alpha$ = 0 and reduces to Poisson
  
## Negative binomial regression

```{r}
library(MASS)
m3 <- glm.nb(data = dat, 
             alcohol ~ sensationC)
summary(m3)
```

## Compare Poisson to negative binomial

:::: {.columns}

::: {.column width="50%"}

```{r}
summary(m1)
```

:::

::: {.column width="50%"}

```{r}
summary(m3)
```

:::

::::

## Compare all three models

* Poisson regression ($\psi$ = 1, $\alpha$ = 0)

```{r}
#| echo: false
library(broom)
m1_table <- as.data.frame(tidy(m1))
m1_table
```

* Overdispersed Poisson regression ($\psi$ = 2.847, $\alpha$ = 0)

```{r}
#| echo: false
m2_table <- as.data.frame(tidy(m2))
m2_table
```

* Negative binomial regression ($\theta$ = 1.393, $\alpha$ = $1/\theta$ = 0.718)

```{r}
#| echo: false
m3_table <- as.data.frame(tidy(m3))
m3_table
```

## Plot: Poisson vs negative binomial

```{r}
#| code-fold: true
nb_b0 <- summary(m3)$coefficients[1]
nb_b1 <- summary(m3)$coefficients[2]
ggplot(data = dat, aes(x = sensationC, y = alcohol)) +
    geom_point(size = 3, 
               alpha = 0.4, 
               color = "#3697aa") +
    labs(x = "Sensation seeking (centered)", y = "Number of alcoholic drinks") +
    geom_smooth(method = "glm", 
                method.args = list(family = 'poisson'), 
                se = FALSE, 
                color = "black", 
                fullrange = TRUE) +
    geom_function(fun = function(x) exp(nb_b0 + nb_b1 * x),
                color = "#dc1e34", 
                size = 1)
```

## Observed vs predicted

```{r}
#| code-fold: true
ggplot(data = dat, aes(x = alcohol)) + 
    geom_bar(fill = "#3697aa", 
             color = "#76777a", 
             size = 1) +
labs(x = "Number of alcoholic drinks", y = "Frequency") +
ylim(0, 90)
```

## Observed vs predicted

```{r}
#| code-fold: true
library(pscl)       # predprob() function for predicted counts
                    # zeroinfl() function for zero-inflated count models
                    # hurdle() for hurdle count models
                    # vuong() for non-nested model comparisons
phat.poi <- predprob(m1)
phat.poi.mn <- apply(phat.poi, 2, mean)
#phat.poi.mn
phat.nb <- predprob(m3)
phat.nb.mn <- apply(phat.nb, 2, mean)
#phat.nb.mn
#phat.zinb <- predprob(zinb)
#phat.zinb.mn <- apply(phat.zinb, 2, mean)
#phat.zinb.mn
x <- seq(0, 17, 1)
probs <- data.frame(x, phat.poi.mn, phat.nb.mn)#, phat.zinb.mn)
#probs
ggplot(data = dat, aes(x = alcohol)) + 
    geom_bar(fill = "#3697aa", 
             color = "#76777a", 
             size = 1) +
    geom_line(data = probs, 
              aes(x = x, y = 400*phat.poi.mn), 
              linetype = "solid",
              size = 1,
              color = "black") +
labs(x = "Number of alcoholic drinks", y = "Frequency") +
ylim(0, 90)
```


## Observed vs predicted

```{r}
#| code-fold: true
ggplot(data = dat, aes(x = alcohol)) + 
    geom_bar(fill = "#3697aa", 
             color = "#76777a", 
             size = 1) +
    geom_line(data = probs, 
              aes(x = x, y = 400*phat.poi.mn), 
              linetype = "solid",
              size = 1,
              color = "black") +
    geom_line(data = probs, 
              aes(x = x, y = 400*phat.nb.mn), 
              size = 1,
              #linetype = "dashed",
              color = "#dc1e34") +
labs(x = "Number of alcoholic drinks", y = "Frequency") +
ylim(0, 90)
```

## Which model to pick?

* **Poisson regression**
  * You will likely never satisfy the equidispersion assumption
* **Overdispersed Poisson regression**
  * Preferred to Poisson regression
  * Appropriate standard errors and relatively simple model
* **Negative binomial regression**
  * More complex, but better able to reproduce the observed counts
  * This may be preferred if you’re interested in prediction

## A word of caution...

Berk & MacDonald (2008)

> If apparent overdispersion results from **specification errors in the systematic part** of the Poisson regression model, resorting to the negative binomial distribution does not help. It can make things worse by giving a **false sense of security** when the fundamental errors in the model remain.

* In other words, negative binomial regression doesn’t *fix a bad model*


# Zeroes

## Zeroes can be really important

* Conceptually, zeroes are meaningful

- Lowest possible value of a count
- Indicate "nothing"
- Two situations:
  - Too few zeroes (relative to a Poisson distribution)
  - Too many zeroes (relative to a Poisson distribution)

## Too few zeroes 

* Sometimes you have a study where the outcome is a count, but it **cannot take on a value of 0**
  * Study of medical visits 
    * **Have to visit the doctor** to get involved in the study
  * Study of substance use 
    * Only interested in **substance users**
* **Truncated Poisson regression** model uses a *truncated Poisson distribution* that removes the probability of zeroes

## Excess zeroes

* Counts often display **"excess" zeroes**
  * More values of 0 than expected for a Poisson distribution
* Even if the rest of the distribution is approximately Poisson, the "excess" zeroes lead to *overdispersion*
  * Sometimes, what looks like overdispersion is really excess zeroes
* There are specific Poisson family models to appropriately deal with excess zeroes, depending on *why* the zeroes are there

## Why are there all these zeroes?

* This is a substantive question
  * You need to know about the outcome you’re studying
* Question: Do the **people who are responding zero** have **some probability** of responding otherwise?

## Some probability of responding > 0?

* Yes $\rightarrow$ **zero-inflated Poisson regression**
  - Cigarettes smoked for smoker who hasn't smoked yet **today**
  - Alcohol consumed for someone who hasn't had any yet **today**
* No $\rightarrow$ **hurdle regression** or **with-zeroes regression**
  - Cigarettes smoked: nonsmoker always responds 0
  - Alcoholic drinks consumed: abstainer always responds 0
  - (These are called "structural zeroes")

## Zero-inflated Poisson regression

* There are zeroes have **some probability** to be non-zero
* Two parts modeled simultaneously:
  - Logistic regression: *Is each zero structural or not?*
  - Poisson regression: non-structural zeroes and positive values
* Can use same set of predictors in both parts, but do not have to
* Can also **model non-structural zeroes and positive values** using *overdispersed* Poisson regression or *negative binomial* regression

## Hurdle regression (with-zeroes) 

* There are zeroes that have **no probability** of being non-zero
  - Two different populations: smokers and nonsmokers, drinkers and nondrinkers
* Two parts modeled simultaneously:
  - Logistic regression: *whether zero or not*
  - Truncated Poisson regression: positive values only
* Can use same set of predictors in both parts, but do not have to

# Variable lengths of time

## Variable lengths of time

* Poisson regression and variants assume a *fixed length of time*
  * Number of aggressive acts committed by a child in 1 hour
  * Number of visits to doctor in past 6 months
  * Number of adverse events in a study
* Often, we measure a count over some *variable period of time*

## Variable lengths of time

- Can extend Poisson type models to variable time periods
  - Include natural log of the measurement interval as a predictor with regression coefficient equal to 1:

$$ln(\hat{\mu}) = ln(time) + b_0 + b_1X_1 + b_2X_2 + \cdots + b_pX_p$$

- In software: include time (specifically, ln(time)) as an "offset"

# Evaluation and comparison

## Deviance

* As with logistic regression
  * Estimated with **maximum likelihood**
  * Use **deviance** to estimate pseudo-$R^2$
* Deviance is a function of the log-likelihood
  * How far model is from a *theoretical* **perfect model**
  * **Relative measure** so can only **compare** to another model
* Deviance usually = -2LL, but **not** for Poisson models

## Pseudo $R^2$

$$R^2_{deviance} = 1 - \frac{deviance_{model}}{deviance_{intercept\ only}}$$

* Deviance for your model is "Residual deviance" from output
* Deviance for intercept model is "Null deviance" from output
* Theoretically ranges from 0 to 1 (but not always in practice)

## Compare models

* Compare **deviances** of 2 **nested models** using likelihood ratio test
* e.g., compare `m1` (only `sensationC`) to `m1add` (`sensationC` and `sex`)

```{r}
anova(m1, m1add, test = "LRT")
```

## Nested models

* Two models are **nested** if 
  * All terms in one are **included in the other**
  * You can get from one to the other by **fixing some paths to a value**
* Why do we care?
  * Likelihood ratio tests can only be used to compare *nested models*
    * If your models are not nested, you **cannot** use the likelihood ratio tests to compare them
  * Instead, use **AIC**, **BIC** (smaller is better), Vuong test

## Nested models 

1. Linear regression with predictor $X_1$
2. Linear regression with predictors $X_1$ and $X_2$

- Model 1 is nested within model 2
  * If the coefficient for $X_2$ = 0 then Model 2 becomes Model 1

## Complicated nesting situations 1

3. Poisson regression with predictors $X_3$ and $X_4$
4. Overdispersed Poisson regression w/ predictors $X_3$ and $X_4$

- Model 3 seems like it's nested within model 4
  * But it's not really
- Same number of things estimated: df = 0

## Complicated nesting situations 2

5. Overdispersed Poisson regression w/ predictor $X_5$
6. Overdispersed Poisson regression w/ predictors $X_5$ and $X_6$

- Model 5 is NOT nested within model 6
- In addition to the extra predictor ($X_6$), the *overdispersion parameter* is not the same in these models

  - Model 5 has a $\psi$ parameter based on only $X_5$
  - Model 6 has a **different** $\psi$ parameter based on both $X_5$ and $X_6$
  - Can't get from Model 6 to Model 5 by setting coefficient to 0

## Complicated nesting situations 3

7. Overdispersed Poisson regression w/ predictors $X_7$ and $X_8$
8. Negative binomial regression w/ predictors $X_7$ and $X_8$

- Model 7 is NOT nested within model 8 (or vice versa)
- The overdispersion parameters are **different** for the two models

  - Model 7 has a $\psi$ parameter and estimates variance = $\psi \mu$ 
  - Model 8 has a $\alpha$ parameter and estimates variance = $\mu + \alpha \mu^2$

## Vuong test

* For nested or non-nested models
* `vuong()` function in **pscl** package

```{r}
library(pscl)
vuong(m1, m3)
```

# Summary

## Summary

* When your outcome is a **count**, use a model from the **Poisson regression** family
  * *Nonlinear, exponential model* or *log-linear model*
  * *Overdispersion*: overdispersed Poisson or negative binomial
  * Too many or too few *zeroes*
  * Variable lengths of time

## In class

* Run some Poisson regression models
  * Poisson
  * Overdispersed Poisson
  * Negative binomial
* Interpret the results

## Next time

* ~~Logistic regression for **binary** outcomes~~
* ~~Poisson regression for **count** outcomes~~
* Survival analysis for **survival** outcomes

