---
title: "Introduction to Biostatistics"
format:
  # html: 
  #   toc: true
  #   toc-location: left
  #   slide-level: 3
  #   number-sections: true
  #   toc-depth: 2
  #   self-contained: true
  #   #output-file: "Mult22-1-handout"
  # pdf: 
  #   toc: true
  #   number-sections: true
  #   #slide-level: 3
  #   toc-depth: 2
  #   shift-heading-level-by: 0
  #   number-offset: [2]
  revealjs:
    toc: false
    number-sections: true
    #slide-level: 3
    scrollable: true
    embed-resources: true
    slide-number: true
    width: 1200
knitr: 
  opts_chunk:
    echo: true
    message: false
    warning: false
---

```{r, include = FALSE, echo = FALSE, wanring = FALSE, message= FALSE}
library(tidyverse)
theme_set(theme_classic(base_size = 16))
set.seed(12345)
```

# Learning objectives

## Learning objectives

* **Add predictors** to a linear regression model
* Interpret **partial regression coefficients**
* **Re-code predictors** to meet assumptions
* **Compare** different models using $R^2_{change}$

# Adding predictors

## Review: Linear regression

* Linear regression
  * **One** *continuous* outcome variable
  * **One or more** predictor variable

## Bivariate regression

* **One predictor**, one outcome
  * **Slope** of line relating predictor and outcome
    * Expected change in $Y$ for 1 unit change in $X$
  * If predictor and outcome are **standardized**
    * Regression coefficient = correlation between $X$ and $Y$

## Multiple regression

* With multiple predictors
  * **Correlations between predictors** affect regression coefficients
  * Not just how $X$ relates to $Y$
    * **How does $X$ relate to other predictors?**
* Interpretation changes
  * Subtle but important

## Correlations and partial correlations

* "Correlation" (bivariate correlation)
  * **Ignore** the effect of other predictors
* Partial correlation
  * **"Control for"** or **"partial out"** the effect of other predictors 
  * Other variables can be related to $X$, $Y$, or both

## Three variables

Three variables that are all related to each other

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (1.5,2.5) circle (2cm) node {$X_2$};
\end{tikzpicture}
```

## Bivariate correlation

Correlation between $X_1$ and $Y$, ignoring $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (1.5,2.5) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (1.5,2.5) circle (2cm);
\end{tikzpicture}
```

## Partial correlation

Correlation between $X_1$ and $Y$, removing the effect of $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (1.5,2.5) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\begin{scope}
\clip (0,0) circle (2cm);
\clip (3,0) circle (2cm);
\fill[white] (1.5,2.5) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (1.5,2.5) circle (2cm);
\end{tikzpicture}
```

## Uncorrelated predictors

Correlation between $X_1$ and $Y$, no effect of $X_2$

```{tikz, echo = FALSE, fig.align="center", fig.height = 3.5, fig.width = 5, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{shapes}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=1cm,y=1cm]
\clip(-2.5,-2.5) rectangle (5.5,5.5);
\draw [line width=2pt] (0,0) circle (2cm) node {$X_1$};
\draw [line width=2pt] (3,0) circle (2cm) node {$Y$};
\draw [line width=2pt] (3,3) circle (2cm) node {$X_2$};
\begin{scope}
\clip (0,0) circle (2cm);
\fill[blue] (3,0) circle (2cm);
\end{scope}
\begin{scope}
\clip (0,0) circle (2cm);
\clip (3,0) circle (2cm);
\end{scope}
\draw [line width=2pt] (0,0) circle (2cm);
\draw [line width=2pt] (3,0) circle (2cm);
\draw [line width=2pt] (3,3) circle (2cm);
\end{tikzpicture}
```

## Example data

* `ICU` data from the **Stat2Data** package

```{r}
library(Stat2Data)
data(ICU)
head(ICU)
```

## Correlations between variables

```{r}
ICU_sub <- ICU[, c(3, 8, 9)]
cor(ICU_sub)
```


## Three models

* `Pulse ~ Age`

```{r}
pulse_vs_age <- lm(data = ICU, 
                   Pulse ~ Age)
```

* `Pulse ~ Emergency`

```{r}
pulse_vs_ER <- lm(data = ICU, 
                  Pulse ~ Emergency)
```

* `Pulse ~ Age + Emergency`

```{r}
pulse_vs_age_ER <- lm(data = ICU, 
                      Pulse ~ Age + Emergency)
```

## Compare models

```{r}
library(modelsummary)
modelsummary(list(pulse_vs_age, pulse_vs_ER, pulse_vs_age_ER))
```

## Just the two predictor model

```{r}
summary(pulse_vs_age_ER)
```

## Interpretation

* $\hat{Pulse} = 84.913 + 0.097 * Age + 11.452 * Emergency$

* The intercept ($b_0$) was 84.913, $t(1)$ = 11.991, $p$ < .0001. An individual who was *0 years old* **and** had a *non-emergency admission* is expected to have a pulse rate of 84.913. 
  * $\hat{Pulse} = 84.913 + 0.097 * 0 + 11.452 * 0 = 84.913$

## Interpretation

* The slope for age ($b_1$) was 0.097, $t(1)$ = 1.021, $p$ > .05. For each 1 year increase in age, the expected pulse rate increased by 0.097 points, **holding admission type constant**; this effect was not significant.

* The slope for admission type ($b_2$) was 11.452, $t(1)$ = 2.652, $p$ < .01. The difference in expected pulse rate between non-emergency admissions (0) and emergency admissions (1) was 11.452 points, **holding age constant**.

## Interpretation

* The $R^2$ for this model was 0.036, F(2, 197) = 3.659, p < .05. About 3.6% of the variance in pulse rate was **explained by** *age* **and** *admission type*.
  * $F$-test for $R^2_{multiple}$ has two different degrees of freedom
  * Numerator df: # of predictors = 2
  * Denominator df: $N$ - # of predictors - 1 = 200 - 2 - 1 = 197

## Two regression lines

* Simplify regression equation for each `Emergency` group
  * `Emergency` = 0
    * $\hat{Pulse} = 84.913 + 0.097 * Age + 11.452 * 0$
    * $\hat{Pulse} = 84.913 + 0.097 * Age$
  * `Emergency` = 1
    * $\hat{Pulse} = 84.913 + 0.097 * Age + 11.452 * 1$
    * $\hat{Pulse} = 96.365 + 0.097 * Age$

## Two parallel lines

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU,
       aes(x = Age, 
           y = Pulse, 
           color = as.factor(Emergency))) +
  geom_point(alpha = 0.4,
             size = 2) +
  geom_abline(intercept = 84.913, 
              slope = 0.097, 
              color="#F8766D",
              size = 1) +
  geom_abline(intercept = 96.365, 
              slope = 0.097, 
              color="#00BA38",
              size = 1)
```

* Same slope for each line: 0.097
* Same difference between lines at any `Age`: 11.452

## Check assumptions

* With more than 1 predictor
  * Residual vs predictor 1
  * Residual vs predictor 2
  * Residual vs predicted outcome value (combination of predictor 1 and 2)

```{r}
ICU$pred <- fitted(pulse_vs_age_ER)
ICU$resi <- residuals(pulse_vs_age_ER)
```

## Residual vs `Age`

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Age, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Residual vs `Emergency`

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = Emergency, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Residual vs predicted `Pulse`

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = pred, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Distribution of residuals

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = resi)) + 
  geom_histogram()
```


# Coding predictors

## Binary predictors

* Simplest coding scheme is 0,1
  * Intercept = expected outcome value for group coded zero
* Other options
  * 1,2 or -1,+1
  * Intercept is less interpretable

## 3 or more categories

* Dummy codes: Extension of 0,1 scheme to more categories
  * Also called "indicator variables"
* Note: 2 categories $\rightarrow$ 1 indicator variable
  * Always use $a - 1$ variables for $a$ categories

## 3 or more categories

* `AgeGroup` in the `ICU` data frame has **3 categories**
  * 1 = young (under 50)
  * 2 = middle (50-69)
  * 3 = old (70+)
* Create 3 indicator variables (dummy codes)
  * Value of 1 for a category, 0s for **all other** categories
    
## 3 or more categories

```{r}
#| code-fold: true
ICU <- ICU %>% mutate(d1 = ifelse(AgeGroup == 1, 1, 0),
                      d2 = ifelse(AgeGroup == 2, 1, 0),
                      d3 = ifelse(AgeGroup == 3, 1, 0))

head(ICU)
```

## Analysis

* Select **any two** (in general, $a - 1$) dummy codes as predictors
  * The one that's left out is the reference group (all 0s)
  * **All other groups are compared to the reference group**
  
AgeGroup | d1 | d2 | d3
--------|--------|--------|--------
1 | 1 | 0 | 0
2 | 0 | 1 | 0
3 | 0 | 0 | 1

* Leave out `d3`: People in `AgeGroup` = 3 have 0s on both `d1` and `d2`

## Analysis

```{r}
pulse_vs_agegroup <- lm(data = ICU, Pulse ~ d1 + d2)
summary(pulse_vs_agegroup)
```

## Interpretation

* **Intercept**: Expected `Pulse` for reference group (`AgeGroup` = 3)
* **Effect of `d1`**: Expected *difference* between the reference group (`AgeGroup` = 3) and the group coded 1 on `d1` (`AgeGroup` = 1)
* **Effect of `d2`**: Expected *difference* between the reference group (`AgeGroup` = 3) and the group coded 1 on `d2` (`AgeGroup` = 2)

## Interpretation

* `Pulse` for `AgeGroup` = 3 is the **intercept**
  * 99.359
* `Pulse` for `AgeGroup` = 1 is 1.563 points **lower** (negative) than the value for `AgeGroup` = 3
  * 99.359 - 1.563 = 97.796
* `Pulse` for `AgeGroup` = 2 is 0.069 points **higher** (positive) than the value for `AgeGroup` = 3
  * 99.359 + 0.069 = 99.428

## Interpretation: Figure

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU,
       aes(x = AgeGroup, y = Pulse)) +
  geom_point() +
  geom_hline(yintercept = 99.359,
             color = "black") + #3
  geom_hline(yintercept = 97.796,
             color = "blue") + #1
  geom_hline(yintercept = 99.428,
             color = "red",
             linetype = "dashed") + #2
  coord_flip()
```

# Centering predictors

## What is an intercept?

* The **expected value** of the *outcome* when **all** predictors are *equal to 0*
  * (We just saw how this worked for dummy codes)
  * 0 may not exist for some predictors
  * 0 may be a non-meaningful value
* How can we make the 0 value more meaningful?

## Centering predictors

* Move the 0 point to a new value
* Typically **mean centering**
  * But you can choose other (meaningful) values
  * Median, percentile, cut-off, etc.
* Subtract this value (e.g., mean) from all predictor values
* Mean of original variable = 0 in new variable
  * 0 is now a meaningful value, so the intercept is **interpretable**
  
## Example: Mean center `Age`

```{r}
mean(ICU$Age, na.rm = TRUE)
ICU$AgeC = ICU$Age - mean(ICU$Age, na.rm = TRUE)
head(ICU)
```

## Model with centered predictor

```{r}
pulse_vs_ageC <- lm(data = ICU, Pulse ~ AgeC)
summary(pulse_vs_ageC)
```

## Model with centered predictor

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = ICU, 
       aes(x = AgeC, y = Pulse)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE) +
  annotate("pointrange", x = 0, 
           y = 98.925, 
           ymin = 98.925, 
           ymax = 98.925,
           colour = "red", 
           size = 2)
```

* $b_0$ = 98.925: Predicted `Pulse` at **mean** `Age`

## Other implications

* Centering predictors doesn't change the **slopes**, only intercepts
* Centering predictors doesn't change the $R^2_{multiple}$
* Interactions / moderation:
  * Conditional effects: Effect of one predictor **depends on** other predictors
  * Specifically, conditional on predictors being equal to 0
    * We'll care where 0 is

# Transforming predictors

## "Linear in the predictors"

* $\hat{Y} = b_0 + b_1 X_1 + b_2 X_2 + \cdots$
  * Predicted value is a **linear combination** of *regression coefficients* and *predictors*
* In practice, this **looks like** a linear relationship between $X$ and $Y$
  * What if there's not a straight line relationship?
  
## Gapminder data

```{r}
library(gapminder)
data(gapminder)
ggplot(data = gapminder,
       aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth()
```

## Bad model

```{r}
bad_gdp <- lm(data = gapminder, lifeExp ~ gdpPercap)
gapminder$bad_pred <- fitted(bad_gdp)
gapminder$bad_resi <- residuals(bad_gdp)
summary(bad_gdp)
```

## Bad model: Fitted line

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm")
```


## Bad model: Residual vs predictor

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = gdpPercap, y = bad_resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Bad model: Residual distribution

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = bad_resi)) + 
  geom_histogram()
```


## Transforming predictors

* It's generally ok to transform **predictors** to make this work
  * Log transformation for very skewed predictors (e.g., GDP, income)
  * Add squared term ($X^2$) for curvilinear effects
* How do you know what transformation to use?
  * Literature
  * Plots
* Note: **Do not** transform outcomes

## Example: Transform predictors

```{r}
gapminder <- gapminder %>% mutate(loggdpPercap = log(gdpPercap))
ggplot(data = gapminder,
       aes(x = loggdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth()
```

## Better model

```{r}
good_gdp <- lm(data = gapminder, lifeExp ~ loggdpPercap)
gapminder$good_pred <- fitted(good_gdp)
gapminder$good_resi <- residuals(good_gdp)
summary(good_gdp)
```

## Better model: Fitted line

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = loggdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth(method = "lm")
```


## Better model: Residual vs predictor

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = loggdpPercap, y = good_resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()
```

## Better model: Residual distribution

```{r}
#| output-location: column
#| results: hold
#| echo: true
ggplot(data = gapminder, 
       aes(x = good_resi)) + 
  geom_histogram()
```


# Comparing models

## Comparing models

```{r}
modelsummary(list(pulse_vs_age, pulse_vs_ER, pulse_vs_age_ER))
```

## $R^2_{change}$

* Change in $R^2$ between models
  * Test whether that change is significant
  * Is more predictors worth a more complex model?
* Only for **nested models**
  * Same model, just add a new predictor
  * Can't compare e.g., untransformed to transformed model

## Test of $R^2_{change}$

```{r}
anova(pulse_vs_age, pulse_vs_age_ER)
```

* This test is **significant**, so
  * **Yes**, adding `Emergency` to a model with `Age` significantly increases the $R^2$ for the model from .001 to .036; F(1, 197) = 7.0326, p<.01

## Test of $R^2_{change}$

```{r}
anova(pulse_vs_ER, pulse_vs_age_ER)
```

* This test is **not significant**, so
  * **No**, adding `Age` to a model with `Emergency` does not significantly increase the $R^2$ for the model from .031 to .036; F(1, 197) = 1.0415, p>.05

## AIC and BIC

* Use AIC or BIC to compare **non-nested models**
  * No statistical test: **Smaller is better**
  * Penalizes more complex models (i.e., more predictors)
* `AIC()` or `BIC()` functions from **stats** package (always loads)

```{r}
AIC(pulse_vs_age)
AIC(pulse_vs_ER)
AIC(pulse_vs_age_ER)
```


## AIC for transformed models

```{r}
AIC(bad_gdp)
AIC(good_gdp)
```

```{r}
BIC(bad_gdp)
BIC(good_gdp)
```

# Next time

## Expanding the model (more)

* Hypothesis testing
* More checking assumptions: Residuals and outliers
* Estimation of the model
  * $R^2$, likelihood
* Interaction (moderator) effects

