---
title: "BTS 510 Lab 8"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Assess** models for **(multi)collinearity** among predictors
* **Conduct** outlier analyses to determine extreme and/or problematic cases

## Data 

* `FirstYearGPA` data from the **Stat2Data** package: $n$ = 219 subjects
  * `GPA`: First-year college GPA on a 0.0 to 4.0 scale
  * `HSGPA`: High school GPA on a 0.0 to 4.0 scale
  * `SATV`: Verbal/critical reading SAT score
  * `SATM`: Math SAT score
  * `Male`: 1= male, 0= female
  * `HU`: Number of credit hours earned in humanities courses in high school
  * `SS`: Number of credit hours earned in social science courses in high school
  * `FirstGen`: 1= student is the first in her or his family to attend college, 0=otherwise
  * `White`: 1= white students, 0= others
  * `CollegeBound`: 1=attended a high school where >=50% students intended to go on to college, 0=otherwise

## Research question(s)

* How do **all these variables** impact first year GPA (`GPA`)?
  * Are there any problems with **collinearity** among the *predictors*?
  * Are there **problematic cases** that are *influencing the results*?

## Tasks

1. Run a **linear regression model** predicting `GPA` from *all other variables* in the dataset.

```{r}
library(Stat2Data)
data(FirstYearGPA)
FirstYearGPA <- FirstYearGPA %>% 
  mutate(ID = row_number())
head(FirstYearGPA)
```

```{r}
m1 <- lm(data = FirstYearGPA, GPA ~ . - ID)
summary(m1)
```

2. Some of those variables seem like they could be **strongly related**, which could cause problems for the model. Check to see if **collinearity** is an issue for this model. To check this, look at:
  * **Correlations among variables**
  * **Variable inflation factors (VIFs)**
  
```{r}
cor(FirstYearGPA)
library(car)
vif(m1)
```
  
3. Based on the **correlations between each predictor and the outcome** (above), are there predictors that you think **should** be significant but aren't? Why do you think they aren't? (This is a bit of a philosophical question.)

4. Are there **extreme cases** that we might be concerned about? Check for extreme values in terms of:
  * **Predictors**
  * **Predicted values**
  * **Changes to predicted values**
  
```{r}
library(broom)
m1_aug <- augment(m1)
m1_aug

ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .hat)) +
  geom_point() +
  geom_hline(yintercept = 3*(9+1)/219, 
             color = "red",
             linetype = "dashed") +
  geom_hline(yintercept = 2*(9+1)/219,
             color = "blue") +
  geom_text(aes(label=ifelse((.hat > 2*(9+1)/219), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```
  
```{r}
library(car)
m1_stresid <- FirstYearGPA %>% mutate(esresid = rstudent(m1))
head(m1_stresid)

ggplot(data = m1_stresid,
       aes(x = c(1:nrow(m1_stresid)), 
           y = esresid)) +
  geom_point() +
  geom_hline(yintercept = 2, 
             color = "blue",
             linetype = "dashed") +
  geom_hline(yintercept = -2,
             color = "blue",
             linetype = "dashed") +
  geom_text(aes(label=ifelse((esresid > 2), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2) +
  geom_text(aes(label=ifelse((esresid < -2), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```

```{r}
ggplot(data = m1_aug,
       aes(x = c(1:nrow(m1_aug)), 
           y = .cooksd)) +
  geom_point() +
  geom_hline(yintercept = 1, 
             color = "red",
             linetype = "dashed") +
  geom_text(aes(label=ifelse((.cooksd > 1), 
                             as.character(ID), 
                             '')),
            hjust = 0, nudge_x = 2)
```
  
5. **Summarize** your findings about collinearity and outliers. *Use plain language.*

* Leverage 

```{r}
m1_aug %>% filter(.hat > 2*(9+1)/219) %>% select(ID)
```

* Discrepancy

```{r}
m1_stresid %>% filter(esresid > 2) %>% select(ID)
m1_stresid %>% filter(esresid < -2) %>% select(ID)
```

* Influence

```{r}
m1_aug %>% filter(.cooksd > 1) %>% select(ID)
```

13 variables had high leverage as measured by leverage measure from the hat matrix with a cut-off of 2k/n. An additional 7 had high discrepancy as measured by the externally studentized residuals with a cut-off of $\pm$ 2.

There were no cases with high influence as measured by Cook's D with a cut-off of greater than 1. This indicates that no cases were changing the overall predicted values.
