---
title: "BTS 510 Lab 7"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Add predictors** to a linear regression model
* Interpret **partial regression coefficients**
* **Re-code predictors** to answer your research questions
* **Compare** different models using $R^2_{change}$

## Data 

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

## Research question

**How do infection status (`Infection`) and admission type (`Emergency`) predict blood pressure?**

## Tasks

```{r}
library(Stat2Data)
data(ICU)
```

1. Conduct **three linear regression models** to address the above research questions.

* Model 1: Infection status predicts blood pressure
* Model 2: Admission type predicts blood pressure
* Model 3: Infection status *and* admission type predict blood pressure

```{r}
m1 <- lm(data = ICU, SysBP ~ Infection)
m2 <- lm(data = ICU, SysBP ~ Emergency)
m3 <- lm(data = ICU, SysBP ~ Infection + Emergency)
```

```{r}
summary(m1)
summary(m2)
summary(m3)
library(modelsummary)
modelsummary(list(m1, m2, m3))
```

```{r}
cor(ICU[, c(6, 7, 9)])
```

2. **Which is the best model?** 

* Use the $F$-test for $R^2_{change}$ and the AIC values to help you decide. Report the tests and values that you used to decide this. 

```{r}
anova(m3, m1)
anova(m3, m2)
AIC(m1)
AIC(m2)
AIC(m3)
```

3. Report the results of the **best** model, including:
  * Intercept value, test statistic, $p$-value, interpretation (in words)
  * Slope value(s), test statistic(s), $p$-value(s), interpretation(s) (in words)
  * $R^2$ value, test statistic, $p$-value, interpretation (in words)

```{r}
m3_coeff <- summary(m3)$coefficients
m3_coeff
```

::: {.callout-note}
Look at the code for the section below to see how to programmatically (i.e., automatically) grab the coefficient values from the output and have them print in your document. 
:::

* $\hat{SysBP} = `r round(m3_coeff[1], 3)` + `r round(m3_coeff[2], 3)` \times Infection + `r round(m3_coeff[3], 3)` \times Emergency$

* The intercept ($b_0$) was `r round(m3_coeff[1], 3)`, t(1) = 31.995, p < .0001. An individual who did not have an infection and had a non-emergency admission is expected to have blood pressure of `r round(m3_coeff[1], 3)`.

* The slope for `Infection` ($b_1$) was `r round(m3_coeff[2], 3)`, t(1) = -2.927, p<.01. Holding `Emergency` constant, the difference between in blood pressure between an infection and a non-infection patient is `r round(m3_coeff[2], 3)` points. Patients with an infection have lower blood pressure.

* The slope for `Emergency` ($b_2$) was `r round(m3_coeff[3], 3)`, t(1) = -2.160, p<.05. Holding `Infection` constant, the difference between in blood pressure between an emergency admit and a non-emergency admit patient is `r round(m3_coeff[3], 3)` points. Patients with an emergency admission have lower blood pressure.

4. Based on **model 3**, what are the predicted blood pressures for the 4 combinations of infection status and admission type? (i.e., infected non-emergency, non-infected non-emergency, infected emergency, non-infected emergency)

* Non-infected non-emergency

```{r}
summary(m3)$coefficients[1] + summary(m3)$coefficients[2] * 0 + summary(m3)$coefficients[3] * 0
```

* Infected non-emergency

```{r}
summary(m3)$coefficients[1] + summary(m3)$coefficients[2] * 1 + summary(m3)$coefficients[3] * 0
```

* Non-infected emergency

```{r}
summary(m3)$coefficients[1] + summary(m3)$coefficients[2] * 0 + summary(m3)$coefficients[3] * 1
```

* Infected emergency

```{r}
summary(m3)$coefficients[1] + summary(m3)$coefficients[2] * 1 + summary(m3)$coefficients[3] * 1
```

5. Using the **predicted** and **residual** values, assess whether the assumptions are met for **model 3**. Include the plots and describe your conclusions based on them. 

```{r}
ICU$pred <- fitted(m3)
ICU$resi <- residuals(m3)
```

```{r}
ggplot(data = ICU, 
       aes(x = Infection, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()

ggplot(data = ICU, 
       aes(x = Emergency, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()

ggplot(data = ICU, 
       aes(x = pred, y = resi)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth()

ggplot(data = ICU,
       aes(x = resi)) +
  geom_histogram()
```


6. In plain language, answer the research question: **How do infection status and admission type predict blood pressure?** (This should not be lengthy -- just a couple sentences.)

