---
title: "BTS 510 Lab 9"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Describe** *maximum likelihood estimation* for linear regression 
* **Describe** *hypothesis testing* for linear regression coefficients, including the sampling distribution used and degrees of freedom

## Data 

* `FirstYearGPA` data from the **Stat2Data** package: $n$ = 219 subjects
  * `GPA`: First-year college GPA on a 0.0 to 4.0 scale
  * `HSGPA`: High school GPA on a 0.0 to 4.0 scale
  * `SATV`: Verbal/critical reading SAT score
  * `SATM`: Math SAT score
  * `Male`: 1= male, 0= female
  * `HU`: Number of credit hours earned in humanities courses in high school
  * `SS`: Number of credit hours earned in social science courses in high school
  * `FirstGen`: 1= student is the first in her or his family to attend college, 0=otherwise
  * `White`: 1= white students, 0= others
  * `CollegeBound`: 1=attended a high school where >=50% students intended to go on to college, 0=otherwise

## Tasks

Question 1: How do demographic variables (`Male`, `FirstGen` and `White`) predict first year college GPA (`GPA`)?

Question 2: How does HS GPA (`HSGPA`) predict first year college GPA (`GPA`) over demographic variables (`Male`, `FirstGen` and `White`)?

1. Run the two models above.

```{r}
library(car)
library(Stat2Data)
data(FirstYearGPA)
head(FirstYearGPA)

m1 <- lm(data = FirstYearGPA, GPA ~ Male + FirstGen + White)
summary(m1)
vif(m1)
vcov(m1)

m2 <- lm(data = FirstYearGPA, GPA ~ Male + FirstGen + White + HSGPA)
summary(m2)
vif(m2)
vcov(m2)
```

* Mean center `HSGPA` to make intercept more interpretable

```{r}
FirstYearGPA <- FirstYearGPA %>%
  mutate(HSGPA_C = HSGPA - mean(HSGPA, na.rm = TRUE))
mean(FirstYearGPA$HSGPA, na.rm = TRUE)
m2a <- lm(data = FirstYearGPA, GPA ~ Male + FirstGen + White + HSGPA_C)
summary(m2a)
vif(m2a)
vcov(m2a)
```

2. What are the log-likelihoods of each model? What can you say about the models based on those values? What *can't* you say?

```{r}
logLik(m1)
logLik(m2)
logLik(m2a)
```

* You cannot say anything about the individual likelihoods, only compare them. Likelihood (/log likelihood) are **relative** measures, unlike probability.
* Model 2/2a has a higher (closer to 0) log likelihood than model 1.
* Centering `HSGPA` doesn't change the log likelihood.

3. Compare the two models using a likelihood ratio test (LRT). Report the results of the test. What can you say about the models based on the test?

```{r}
anova(m1, m2, test = "LRT")
-2*(logLik(m1) - logLik(m2))
```

* Model 2 is significantly better than model 1: $\chi^2(1) = 54.1902, p<.0001$.
* Adding `HSGPA` to a model with `Male`, `FirstGen`, and `White` significantly improves the model.

4. Report the results for the better model (based on the LRT). Include all regression coefficients, $R^2$, test statistics, $p$-values.

* Intercept: $b_0 = 2.885, t(214) = 43.321, p<.0001$
  * The expected first year college GPA for a person who is **female**, **not** a  first generation college student, **not** white, and who has mean high school GPA is 2.885.
* `Male`: $b_1 = 0.062, t(214) = 1.144, p>.05$
  * Male students have a first year college GPA that is 0.062 points *higher* than female students, *holding all other predictors constant*. This effect is not significantly different from 0.
* `FirstGen`: $b_2 = -0.185, t(214) = -2.125, p<.05$
  * First generation college students have a first year college GPA that is 0.185 points *lower* than that of non-first generation students, *holding all other predictors constant*; this effect is significant.
* `White`: $b_3 = 0.25738, t(214) = 3.786, p<.001$
  * White students have a first year college GPA that is 0.257 points *higher* than their non-white classmates, *holding all other predictors constant*; this effect is significant.
* `HSGPA_C`: $b_4 = 0.55967, t(214) = 7.751, p<.0001$
  * Each 1 unit increase in high school GPA corresponds to a 0.560 point *higher* first year college GPA, *holding all other predictors constant*. This effect is significant.
* $R^2_{multiple} = 0.2884, F(4, 214) = 21.68, p<.0001$
  * The four predictors in this model account for 28.84% of the variance in the outcome, first year college GPA.
  
  