---
title: "BTS 510 Lab 6"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* Describe the logic of **linear regression** 
* Describe the **assumptions** of linear regression
* Briefly **present results** of linear regression
* Use a **regression equation** to summarize a model
* Compare and contrast **observed**, **predicted**, and **residual** values

## Data 

* `ICU` data from the **Stat2Data** package
  * `ID`: Patient ID code
  * `Survive`: 1 = patient survived to discharge or 0 = patient died
  * `Age`: Age (in years)
  * `AgeGroup`: 1 = young (under 50), 2 = middle (50-69), 3 = old (70+)
  * `Sex`: 1 = female or 0 = male
  * `Infection`: 1 = infection suspected or 0 = no infection
  * `SysBP`: Systolic blood pressure (in mm of Hg)
  * `Pulse`: Heart rate (beats per minute)
  * `Emergency`: 1 = emergency admission or 0 = elective admission

## Tasks

1. **Does blood pressure differ between those with a suspected infection and those without?** *Fit a linear regression to answer this question.*

```{r}
library(Stat2Data)
data(ICU)
m1 <- lm(data = ICU, SysBP ~ Infection)
summary(m1)
```

2. Write up the **interpretation** of the model, including:
  * Intercept value, test statistic, $p$-value
  * Intercept interpretation (in words)
  * Slope value, test statistic, $p$-value
  * Slope interpretation (in words)
  * $R^2$ value, test statistic, $p$-value
  * $R^2$ interpretation (in words)

3. What is the **predicted blood pressure** for someone *with a suspected infection*? What is the **predicted blood pressure** for someone *without a suspected infection*?

* Predict function

```{r}
newdat <- data.frame(Infection = c(0,1))
predict(m1, newdat)
```

* Use R as a smart calculator

```{r}
bp_0 <- summary(m1)$coefficients[1] + summary(m1)$coefficients[2] * 0
bp_0
bp_1 <- summary(m1)$coefficients[1] + summary(m1)$coefficients[2] * 1
bp_1
```

* Use R as a dumb calculator

```{r}
bp_0d <- 138.672 - 15.220 * 0
bp_0d
bp_1d <- 138.672 - 15.220 * 1
bp_1d
```

4. Calculate the **predicted** and **residual** values for each person and add them to the original dataset. **Does it looks like the assumptions are satisfied?** Make plots to help you decide:
  * Residual vs predictor
  * Q-Q plot of residuals
  * Histogram of residuals (all together and separately for the two groups)

```{r}
ICU$pred1 <- fitted(m1)
ICU$resi1 <- residuals(m1)

ggplot(data = ICU, 
       aes(x = Infection, y = resi1)) + 
  geom_point(alpha = 0.3, 
             size = 2) +
  geom_smooth(method = "lm")

ggplot(data = ICU,
       aes(sample = resi1)) + 
       stat_qq() +
       stat_qq_line()

ggplot(data = ICU,
       aes(x = resi1)) +
  geom_histogram()

ggplot(data = ICU,
       aes(x = resi1)) +
  geom_histogram() +
  facet_grid(cols = vars(Infection))
```

5. **Does blood pressure differ between those with a suspected infection and those without? How?** (Brief, simple words, no jargon, no statistics.) 

# Plot.

```{r}
int <- summary(m1)$coefficients[1]
slo <- summary(m1)$coefficients[2]
ggplot(data = ICU,
       aes(x = Infection, y = SysBP)) +
  geom_point(alpha = 0.3) +
  annotate("text", 
    x = 0.5, 
    y = 250, 
    label = bquote(~ hat(Y) == .(round(int, 3)) + .(round(slo, 3))~"* Infection"),
    size = 6)
```

