---
title: "BTS 510 Lab 13"
format:
  html:
    embed-resources: true
    self-contained-math: true
    html-math-method: katex
    number-sections: true
    toc: true
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
---

```{r}
#| label: setup
set.seed(12345)
library(tidyverse)
library(Stat2Data)
theme_set(theme_classic(base_size = 16))
```

## Learning objectives

* **Conduct** analysis on time-to-event outcomes
  * Kaplan-Meier curve, proportional hazards regression
* **Describe** issues related to time-to-event analysis
  * Censoring, proportional hazards, time-varying predictors

## Data 

* `cancer` dataset in **survival** package
* n = 228 observations on 10 variables
  * `inst`:	Institution code
  * `time`:	Survival time in **days**
  * `status`:	censoring status 1=censored, 2=dead
  * `age`:	Age in years
  * `sex`:	Male=1 Female=2
  * `ph.ecog`:	ECOG performance score as rated by the physician. 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed <50% of the day, 3= in bed > 50% of the day but not bedbound, 4 = bedbound
  * `ph.karno`:	Karnofsky performance score (bad=0-good=100) rated by physician
  * `pat.karno`:	Karnofsky performance score as rated by patient
  * `meal.cal`:	Calories consumed at meals
  * `wt.loss`:	Weight loss in last six months (pounds)

## Analysis

* Look more carefully at the last analysis from the lecture
  * `sex` and `age` predict survival time (`time` and `status`)

* Load packages and read in data 

```{r}
library(survival)
library(ggsurvfit)
data(cancer)
```

* Look at the model

```{r}
cox2 <- coxph(Surv(time, status) ~ sex + age, data = cancer)
summary(cox2)
```

* Does the proportional hazards assumption hold for this model?

```{r}
cox.zph(cox2)
```

* `age` is continuous but let's chop it up to look at this effect
  * `cut()` divides the variable based on the `breaks` and then calls each category by the names in `labels`
  * 4 values in `breaks` = 3 categories because they're the cut-points, so `-Inf` to 50, 50 to 70, 70 to `Inf`

```{r}
ggplot(data = cancer, aes(x = age)) +
  geom_histogram()

cancer <- cancer %>%
  mutate(age_group = cut(age, 
                        breaks = c(-Inf, 50, 70, Inf),
                        labels = c(1, 2, 3)))
head(cancer)
```

* Kaplan-Meier curves for each effect (`sex` and `age_group`) separately

```{r}
sfit_sex <- survfit(Surv(time, status) ~ sex, data = cancer)
sfit_age <- survfit(Surv(time, status) ~ age_group, data = cancer)

ggsurvfit(sfit_sex) +
  add_confidence_interval() + add_risktable()
ggsurvfit(sfit_age) +
  add_confidence_interval() + add_risktable()
```

## Tasks

* Select another predictor in the dataset
* Run a Cox regression model with that predictor
  * Report hazard ratios and their confidence intervals
    * Interpret the effects
  * Is the proportional hazards assumption satisfied?
    * If not, what are some time-varying covariates you might include?
* Print Kaplan-Meier curves for the predictor
  * If it's continuous, chop it up in some way that makes sense to you
  * Describe what's going on in the plot
